---
title: "Biostatistics Lab - loglinear models"
author: "Waldron Lab"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

# Introduction

For this exercise we, will be using part of the Census dataset and a
prison-admissions dataset from `TheUpshot` from the NYTimes. Take some time to look at both datasets:

1) Upshot incarceration data, https://github.com/TheUpshot/prison-admissions (use 2014 data)
2) Census poverty data, https://www.census.gov/did/www/saipe/data/statecounty/data/2014.html


# Data acquisition

We'll give you some help with data acquisions.

First, the Upshot data on prison admissions:

```{r}
##  download data
library(readr)
library(readxl)

padmissions <- read_csv("https://raw.githubusercontent.com/TheUpshot/prison-admissions/master/county-prison-admissions.csv")
```

What happened in this "parsing error"?  Figure it out, then correct it:

```{r}
padmissions[1796, "admissions2014"] <- 1403
```

Then, the 2014 census data on poverty:

```{r}
## give the file we will download a name
localFile <- "~/Downloads/est14ALL.xls"

## point to the file with URL
censusWebFile <- "https://www.census.gov/did/www/saipe/downloads/estmod14/est14ALL.xls"

## download file to local address
download.file(url = censusWebFile, destfile = localFile)

## Note: If download not working, run this line below
## browseURL(dirname(censusWebFile))
## Select either the TXT or XLS file (first row in table)

census <- read_excel(localFile, skip = 3)
names(census) = make.unique(names(census))
names(census) <- gsub(" |,", "", names(census))
```

```{r}
library(dplyr)
census <- select(census, -contains("90"))
census <- mutate(census, fips = paste0(StateFIPSCode, CountyFIPSCode))
```


# Join the datasets on 5-digit FIPS state/county codes using dplyr

Merge datasets using dplyr. `inner_join()` keeps only records that can be matched on FIPS code.

# Descriptives

Using joined dataset make histograms of incarceration rates, poverty % variables, and income variable.

Make some other descriptive statistics.

# Univariate log-linear Poisson model

Fit a univariate log-linear model with #/10K incarcerated as y and poverty % (all ages) as x.

Make residuals plots.  How is the fit?

Do this for poverty of different age groups. Which is a stronger predictor, ignoring the issues of model fit?

# Multivariate log-linear Poisson model

How collinear are the different poverty variables and the household income variable?

Try fiting and assessing a multivariate GLM using a Poisson error model and log-linear link function, using the poverty and income variables. Comment.
