---
title: "Biostatistics Lab"
author: "Waldron Lab"
date: "`r format(Sys.time(),'%B %d, %Y')`"
output: html_document
---

For this exercise we, will be using part of the Census dataset and a
prison-admissions dataset from `TheUpshot` from the NYTimes. 

# Data acquisition

First, the Upshot data on prison admissions:

```{r}
library(readr)
library(readxl)
padmissions <- read_csv("https://raw.githubusercontent.com/TheUpshot/prison-admissions/master/county-prison-admissions.csv")
```

Correcting a parse failure (error in data) - correct 1,403 to 1403

```{r}
padmissions[1796, "admissions2014"] <- 1403
```

Then, the 2014 census data on poverty:

```{r}
## give the file we will download a name
localFile <- "~/Downloads/est14ALL.xls"

## point to the file with URL
censusWebFile <- "https://www.census.gov/did/www/saipe/downloads/estmod14/est14ALL.xls"

## download file to local address
download.file(url = censusWebFile, destfile = localFile)

## Note: If download not working, run this line below
## browseURL(dirname(censusWebFile))
## Select either the TXT or XLS file (first row in table)

census <- read_excel(localFile, skip = 3)
names(census) = make.unique(names(census))
names(census) <- gsub(" |,", "", names(census))
```

```{r}
library(dplyr)
census <- select(census, -contains("90"))
census <- mutate(census, fips = paste0(StateFIPSCode, CountyFIPSCode))
```


# Join the datasets on 5-digit FIPS state/county codes using dplyr

Merge datasets using dplyr. `inner_join()` keeps only records that can be matched on FIPS code.

```{r}
prisonpov <- dplyr::inner_join(padmissions, census, by = "fips")
```

# Descriptives

Using joined dataset make histograms of incarceration rates, poverty % variables, and income variable.

Example for incarceration rates:

```{r}
library(ggplot2)
ggplot(prisonpov, aes(admitsPer10k2014)) + 
  geom_histogram() + xlab("Prison Admissions per 10K (2014)")
```

# Univariate log-linear Poisson model

Fit a univariate log-linear model with #/10K incarcerated as y and poverty% as x

```{r}
fit1 <- glm(admitsPer10k2014 ~ PovertyPercentAllAges, data=prisonpov, family=poisson("log"))
```

# Multivariate log-linear Poisson model

Fit a multivariate model, do residuals analysis

