---
title: "Biostats 3 Lab 1"
date: "February 20, 2016"
output: html_document
---

## Logistic Regression using mixed models

Dataset source: Raudenbush and Bryk (2002)

Data description: Data on whether or not children had to repeat a grade during their time in primary school (Thailand, 1988)

Dataset name: `Thai spss.SAV`

Data Codebook:

Rep1 (repeated grade 1= Yes, 0=No)
Male (1=male, 0=female)
PPED (pre-primary school education 1= Yes, 0=No).
SES – school SES centered

1. The dataset is in SPSS format we will import it into R. 
__Note:__ Assuming the dataset file location is in your downloads folder.

Set the working directory to `Downloads` depending on system. 

```{r}
if (.Platform$OS.type == "unix") {
    setwd("~/Downloads")
} else if (.Platform$OS.type == "windows") {
    setwd("~/../Downloads")
}
```

Read the dataset and save it to an object named `thai`. 

```{r}
library(haven)
thai <- read_spss("Thai spss.SAV")
```

2. Run `head(thai)` to identify variables as level 1 or level 2 variables. Conduct descriptive analysis using `table` and `prop.table` for categorial variables and the appropriate summary functions for numeric variables. 

Categorical variables: 

```{r}
head(thai)
table(thai$male); prop.table(table(thai$male))
table(thai$pped); prop.table(table(thai$pped))
table(thai$rep1); prop.table(table(thai$rep1))
```

Numeric variables: 
__Note:__ You can run each function individually. 

```{r}
mean(thai$msesc, na.rm = TRUE)
sd(thai$msesc, na.rm = TRUE)
min(thai$msesc, na.rm = TRUE)
max(thai$msesc, na.rm = TRUE)
```

__Note:__ or as a group 

```{r}
myfuns <- list(mean = mean, stdev = sd, min = min, max = max)
sapply(myfuns, function(x){ round(x(na.omit(thai$msesc)), 3)})
```

To easily create summary statistics by group, `dplyr` offers a quick solution. 

Group by gender `male`: 
```{r}
library(dplyr)
thai %>% group_by(male) %>% summarize(n = n(), mean = mean(rep1, na.rm = TRUE), sd = sd(rep1, na.rm = TRUE))
```

Group by pre-primary school education `pped`: 
```{r}
thai %>% group_by(pped) %>% summarize(n = n(), mean = mean(rep1, na.rm = TRUE), sd = sd(rep1, na.rm = TRUE))
```

3. Run empty model using `lme4::lmer`. Write down the corresponding model (level 1 and level 2 equations)

```{r}
if (!require(lme4)){
    install.packages("lme4")
}
library(lme4)
## check for the appropriate reference variable (0 is 0 and 1 is 1)
contrasts(factor(thai$rep1))
mod1 <- glmer(factor(thai$rep1) ~ 1 + (1 | schoolid), family = binomial(link = "logit"), data = thai, control=glmerControl(optimizer="bobyqa"))
summary(mod1)
```

4. Describe the variation between schools in grade repetition.

5. Interpret fixed and random effects in empty model.

6. Now we will add one level 1 predictor, gender. Interpret fixed effects of gender ($\beta_{10}$).
Write down the equations for the model.

```{r}
mod2 <- glmer(factor(thai$rep1) ~ male + (1 | schoolid), family = binomial(link = "logit"), data = thai, control=glmerControl(optimizer="bobyqa"))
```

7. Now we will add the other level 1 predictors (pped) and level 2 variable (school ses centered).

```{r}
mod3 <- glmer(factor(thai$rep1) ~ male + pped + msesc + (1 | schoolid), family = binomial(link = "logit"), data = thai, control=glmerControl(optimizer="bobyqa"))
summary(mod3)
exp(coef(summary(mod3))[-1 ,1])
```

To obtain 95% confidence intervals from the results of the `glmer` function, run the following code: 

```{r}
results <- as.data.frame(coef(summary(mod3)))
confints <- with(results, Estimate + outer(`Std. Error`, c(lower=-1, upper=1)) * sqrt(qchisq(0.95, 1)))
rownames(confints) <- rownames(results)
newresults <- cbind(results, confints)
newresults
```

8. Interpret the coefficients in the model.

9. Run GEE on the same data and interpret

```{r}
if (!suppressMessages(require("gee"))) {
    install.packages("gee")
}
if (!suppressMessages(require("geesmv"))) {
    install.packages("geesmv")
}
library(gee)
## Unstructured correlation structure
formula <- rep1 ~ male + pped + msesc
gee.unstr <- gee(formula, id = schoolid, family = binomial(link = "logit"),
                        data = thai, corstr = "unstructured")
summary(gee.unstr)
```

GEE with MBN variance estimator

```{r}
library(geesmv)
thai2 <- dplyr::arrange(thai[complete.cases(thai), ])
thai2 <- as.data.frame(thai2)
mbn.unstr <- GEE.var.mbn(formula, id = "schoolid", family = binomial, data = thai2)
```