---
title: "Lab 2"
date: "February 16, 2016"
output: html_document
---

Dataset 1: LangScores

Source: Snijders and Boskers (1999)


We are interested in language scores `langPOST`

1. Run a model with only 1 level-2 predictor, the school SES centered.

```{r}
setwd("~/../Downloads/")
library(haven)
sb <- read_sas("SB.sas7bdat")
mean(sb$schoolSES)
sb$ses_center <- scale(sb$schoolSES, scale = FALSE)
```

Check your work and make sure that the mean of the new variable is approximately zero. 

```{r}
round(mean(sb$ses_center), 3)
```

```{r}
library(nlme)
fitmix <- lme(langPOST ~ ses_center, data = sb, random = ~1 |  schoolNR)
summary(fitmix)
VarCorr(fitmix)
```

2. Interpret the intercept. 

3. Interpret the parameter estimate for school SES. 

4. Based on the results from above do you think school SES is an important predictor of language scores?

5. How much of the between group variation in language scores is due to school SES?

6. Now allow slopes to vary as well. 

```{r}
fitmix2 <- lme(langPOST ~ ses_center, data = sb, random = ~ ses_center | schoolNR)
summary(fitmix2)
VarCorr(fitmix2)
```
