---
title: "Biostats 3 Lab 1"
date: "February 20, 2016"
output: html_document
---
2
## Logistic Regression using mixed models

Dataset source: Raudenbush and Bryk (2002)

Data description: Data on whether or not children had to repeat a grade during their time in primary school (Thailand, 1988)

Dataset name: `Thai spss.SAV`

Data Codebook:

Rep1 (repeated grade 1= Yes, 0=No)
Male (1=male, 0=female)
PPED (pre-primary school education 1= Yes, 0=No).
SES – school SES centered

1. The dataset is in SPSS format we will import it into R. 
__Note:__ Assuming the dataset file location is in your downloads folder.

Set the working directory to `Downloads` depending on system. 

```{r}
library(haven)
if (.Platform$OS.type == "unix") {
    setwd("~/Downloads")
} else if (.Platform$OS.type == "windows") {
    setwd("~/../Downloads")
}
```

Read the dataset and save it to an object named `thai`. 

```{r}
thai <- read_spss("Thai spss.SAV")
```

2. Run `head(thai)` to identify variables as level 1 or level 2 variables. Conduct descriptive analysis using `table` and `prop.table` for categorial variables and the appropriate summary functions for numeric variables. 

Categorical variables: 

```{r}
head(thai)
table(thai$male); prop.table(table(thai$male))
table(thai$pped); prop.table(table(thai$pped))
table(thai$rep1); prop.table(table(thai$rep1))
```

Numeric variables: 
__Note:__ You can run each function individually. 

```{r}
mean(thai$msesc, na.rm = TRUE)
sd(thai$msesc, na.rm = TRUE)
min(thai$msesc, na.rm = TRUE)
max(thai$msesc, na.rm = TRUE)
```

__Note:__ or as a group 

```{r}
myfuns <- list(mean = mean, stdev = sd, min = min, max = max)
sapply(myfuns, function(x){ round(x(na.omit(thai$msesc)), 3)})
```

To easily create summary statistics by group, `dplyr` offers a quick solution. 

Group by gender `male`: 
```{r}
library(dplyr)
thai %>% group_by(male) %>% summarize(n = n(), mean = mean(rep1, na.rm = TRUE), sd = sd(rep1, na.rm = TRUE))
```

Group by pre-primary school education `pped`: 
```{r}
thai %>% group_by(pped) %>% summarize(n = n(), mean = mean(rep1, na.rm = TRUE), sd = sd(rep1, na.rm = TRUE))
```

3. Run empty model using `lme4::lmer`. Write down the corresponding model (level 1 and level 2 equations)

```{r}
library(lme4)
mod1 <- lmer(factor(thai$rep1) ~ 1 + (1 | schoolid), family = binomial(link = "logit"), data = thai)
```