---
title: "Biostats 3 Lab 4"
date: "March 1, 2016"
output: html_document
---

## Longitudinal Data Analysis using mixed models

**_Data Source:_** Everitt and Rabe-Hesketh (2001) (dataset was modified for this class)

**_Data Description:_** Double blind placebo controlled study of the use of estrogen given transdermally in the treatment of post-natal depression

* 61 women with major depression that began within 3 months of childbirth and persisted for up to 18 months postnatally
* Allocated randomly to
    + active treatment (n=34; 3 months of transdermal 17 beta-estradiol 200 mg daily alone, then 3 months with added cyclical dydrogesterone 10 mg daily for 12 days each month) or 
    + placebo (n=27; placebo patches and tablets according to same regime)

**_Dataset name:_** `estrogen`

**_Dataset Codebook:_**

* Outcome variable: Edinburgh Postnatal Depression Scale (EPDS) - self-ratings of depressive symptoms: higher scores imply more depressed
* Assessments made on six visits every two-months for a year after treatment began
one assessments were made prior to the start of the treatment (time 0)
* Treatment ( 1-active, 0-placebo)

As usual, set your working directory to the __Downloads__ folder: (platform dependent)
```{r, eval = FALSE}
if(.Platform$OS.type == "unix") {
    setwd("~/Downloads")
} else if (.Platform$OS.type == "windows") {
    setwd("~/../Downloads")
}
```
__Note:__ Dependencies can be installed via the `install.packages` function. See `?install.packages` for details. 

1. Import SPSS dataset into R: 

```{r}
library(haven)
estrogen <- read_spss("estrogen-LONGFORM.sav")
estrogen$Group <- factor(estrogen$Group, labels = c("Placebo", "Active"))
```

2. Run `head` on the dataset. Identify variables as level 1 or level 2 variables. Conduct descriptive analysis: describe dependent variable at each time point and by treatment group and report rates of missing data.

```{r}
suppressPackageStartupMessages(library(dplyr))
head(estrogen, 10)
group_by(estrogen, time) %>% summarize(avg_epds = mean(EPDS, na.rm = TRUE))
group_by(estrogen, time, Group) %>% summarize(avg_epds = mean(EPDS, na.rm = TRUE))
```

3. Generate individual trajectories graphs of EPDS as a function of time overall and within levels of treatment & regression  by group

Plot individual trajectories with `base` graphics: (not so easy to do by group, see `ggplot2` below)

```{r}
interaction.plot(estrogen$time, estrogen$id, estrogen$EPDS, xlab = "Time", ylab = "EPDS", legend = FALSE, col = c(1:10), lwd = 2)
```

Plot individual trajectories by group easily with `ggplot2::facet_grid`

```{r, warning=FALSE}
library(ggplot2)
ggplot(estrogen, aes(x = time, y = EPDS, group = id)) + geom_line()
ggplot(estrogen, aes(x = time, y = EPDS, group = id)) + geom_line() + facet_grid(. ~ Group)
```

Plot regression line by group

```{r, warning=FALSE}
ggplot(estrogen, aes(x = time, y = EPDS)) + facet_grid(. ~ Group) + geom_smooth(method = "lm", se = FALSE)
```

4. Investigate the serial correlation. Use AIC and BIC to compare the models.

First obtain correlation between measurement occasions using the WIDE data format (which you need to import from SPSS)

```{r}
estrogenW <- read_spss("estrogen-WIDEFORM.sav")
cor(estrogenW[-1], use = "pairwise.complete.obs")
```

Unstructured covariance structure

```{r}
suppressPackageStartupMessages(library(nlme))
fitmix1 <- lme(EPDS ~ 1, data = estrogen, random = ~1 |  id,
              na.action = na.omit,
              control = list(msMaxIter=600, msMaxEval=600,
                 sing.tol=1e-20, msVerbose=FALSE))
summary(fitmix1)
```

Compound Symmetry

```{r}
fitmix2 <- lme(EPDS ~ 1, correlation = corCompSymm(),
              data = estrogen, random = ~1 |  id,
              na.action = na.omit,
              control = list(msMaxIter=600, msMaxEval=600,
                 sing.tol=1e-20, msVerbose=FALSE))
summary(fitmix2)
```

Autoregressive process of order 1

```{r}
fitmix3 <- lme(EPDS ~ 1, correlation = corAR1(),
              data = estrogen, random = ~1 |  id,
              na.action = na.omit,
              control = list(msMaxIter=600, msMaxEval=600,
                 sing.tol=1e-20, msVerbose=FALSE))
summary(fitmix3)
```

Get all the AICs together

```{r}
fits <- list(unstr = fitmix1, compsym = fitmix2, ar1 = fitmix3)
AICs <- sapply(fits, function(fit) {summary(fit)$AIC})
AICs
```
 
5. Run an empty model, compute ICC

Calculate the ICC from an `lme` model fit using the following function:

```{r}
ICClme <- function(out){
  varests <- as.numeric(VarCorr(out)[1:2])
  varests[1] / sum(varests)
}
```

Run it on your empty model

```{r}
VarCorr(fitmix1)
ICClme(fitmix1)
```

6. Include time as level 1 predictor. Are there significant individual differences in baseline depression scores?

```{r}
fitmix4 <- lme(EPDS ~ time, data = estrogen, random = ~1 |  id,
              na.action = na.omit, method = "ML",
              control = list(msMaxIter=600, msMaxEval=600, 
                 sing.tol=1e-20, msVerbose=FALSE))
summary(fitmix4)
VarCorr(fitmix4)
intervals(fitmix4)
```
