---
title: "Biostats 3 Lab 1"
date: "February 2, 2016"
output: html_document
---

Dataset 1: LangScores

Source: Snijders and Boskers (1999)


We are interested in language scores `langPOST`

1. Download from BlackBoard
* SB.sas7bdat

__Note:__ This is an RMarkdown file or .Rmd file. You can directly type your answers in this file and when done, click on the "Knit HTML" button at the top (from within RStudio). RMarkdown files are great for reproducibilty as they can include code and text all in one document.

To run code within the RMarkdown file, you can use the shortcut `Ctrl + Alt + C` within the current chunk where the cursor is positioned.
A chunk is delimited by triple backticks. Run only the code inside those backticks and not the backticks themselves. You can also select `Code` > `Run region` > `Run current chunk` from the RStudio menu to run the code inside the current chunk.

You will need to run this chunk yourself. To install packages for the first time, use the `install.packages` function. Go the the line below and run the current chunk so you can install the `haven` package. This package will be used to read the SAS dataset into memory. 

```{r, eval = FALSE}
install.packages("haven")
```

To load a package use the `library` function.

```{r}
library("haven")
```

Now, read in the data. First, make sure that your working directory is the same as the one that __contains the data__. In this example, we use the downloads directory. Depending on your machine, the `Downloads` folder may be located in different subfolders. The following `if else` control structure accounts for that.

```{r, eval = FALSE}
if (.Platform$OS.type == "unix") {
    setwd("~/Downloads")
} else if (.Platform$OS.type == "windows") {
    setwd("~/../Downloads")
}
```

We will now read the data using a function within the `haven` package and assign it to a variable (any name will suffice, let's use `sb`).

```{r}
sb <- read_sas("SB.sas7bdat")
```

Now we are ready to explore the data. 

To look at tabular data, use the `View` function on the object:

```{r, eval = FALSE}
View(sb)
```

You can also get a small view of what the data looks like by using both the `head` and `tail` functions on the object `sb`.

```{r, eval = FALSE}
head(sb)
tail(sb)
```

2. To look at the dimensions of the table and obtain the number of records, use the `dim` function. It will return the number of rows and the number of columns in the table, respectively.

```{r, eval = FALSE}
dim(sb)
```

3. Use the following code to fill-in Table 1, which is a descriptive table of the individual level factors contained in the SB dataset and language scores (langpost) by these categories. 

First the `table` function calculates the number of participants in each factor level. Then the `prop.table` function calculates proportions in each category for a `table`, in this case `table(sb$sex)`. We then multiply by 100 to get a percent. Then the `tapply` calculates mean language scores by each of levels in the factor variables. In this case, we use `sex` and `minority` as factor variables. 

First, the `sex` variable: 

```{r}
table(sb$sex)
prop.table(table(sb$sex))*100
tapply(sb$langPOST, sb$sex, mean)
```

Now we do it for the `Minority` variable: 

```{r}
table(sb$Minority)
prop.table(table(sb$Minority))*100
tapply(sb$langPOST, sb$Minority, mean)
```

#### __Table 1.__ Individual level characteristics

|            | N | % | Mean Language Score |
|------------|---|---|---------------------|
| Sex        |   |   |                     |
| 0 = Female |   |   |                     |
| 1 = Male   |   |   |                     |
| Minority   |   |   |                     |
| 0 = No     |   |   |                     |
| 1 = Yes    |   |   |                     |

_**Note:**_ If you are reading the HTML version of this page, please edit the table in the .Rmd file in RStudio.
Type your answers between the "|" (table column delimiters) and recompile the document by clicking "Knit HTML" at the top.

Run the code to look at mean language scores by school.

The `dplyr` package allows for readable code. Below can be read as, first group your data by `schoolNR` then you summarize it by N, Means, Standard Deviations, Minima and Maxima.

```{r}
suppressPackageStartupMessages(library(dplyr))
group_by(sb, schoolNR) %>% summarize(n = n(), avg_lang = mean(langPOST), sdev = sd(langPOST), min = min(langPOST), max = max(langPOST))
```

An alternative way of doing these summary calculations with just base functions in R: 

```{r}
myfuns <- list(n = length, mean = mean, stdev = sd, min = min, max = max)
dats <- lapply(myfuns, function(x) {tapply(sb$langPOST, sb$schoolNR, x)})
schools <- do.call(cbind, dats)
head(schools)
```


4. Run the following code to see the first 20 observations for the selected variables. 

```{r}
head(sb[, c("langPOST", "sex", "schoolNR", "schoolSES")],20)
```

5. What do you notice about the distribution of language scores in these 20 individuals who are from the same school? 

6. What do you notice about the distribution of schoolses in these 20 individuals from school 1?

7. One assumption of a general linear mixed model with a continuous outcome is that the outcome at level-1 is normally distributed.  Run the following code to check the distribution of the outcome.

```{r}
hist(sb$langPOST, breaks = 20, density = 20, xlab = "POST Language Score", main = "histogram with normal curve", freq = FALSE, ylim = c(0, 0.05))
curve(dnorm(x, mean = mean(sb$langPOST), sd = sd(sb$langPOST)), col = "darkblue" , lwd = 2, add = TRUE)
```

8. Another assumption of a general linear mixed model with a continuous outcome is that the outcome at level-2 is normally distributed.  Run the following code to check the distribution of the outcome at level-2.

```{r}
hist(dats$mean, breaks = 20, density = 20, xlab = "School Means", main = "histogram with normal curve", freq = FALSE, ylim = c(0, 0.12))
curve(dnorm(x, mean = mean(dats$mean), sd = sd(dats$mean)), col = "darkblue" , lwd = 2, add = TRUE)
```

9. Now run an empty model using `nlme::lme` a function from the `nlme` package. 

```{r, message=FALSE}
suppressPackageStartupMessages(library(nlme))
fitmix <- nlme::lme(langPOST ~ 1, data = sb, random = ~1 |  schoolNR)
summary(fitmix)
```

To see the intervals of the estimate(s) use the `intervals` function:

```{r}
intervals(fitmix)
```

Covariance Parameter Estimates in R

```{r}
VarCorr(fitmix)
```

### Shortcut for calculating ICC from an `lme` object:

```{r}
ICClme <- function(out){
  varests <- as.numeric(VarCorr(out)[1:2])
  varests[1] / sum(varests)
}
```

Calculate the ICC for the current `lme` object:

```{r}
ICClme(fitmix)
```

10. How many schools are there in this sample? 

11. What is the maximum number of subjects in a school? 

12. How many fixed effects were estimated? 

13. How many random effects were estimated? 
	
_Using the Covariance Parameter Estimates portion of the output answer the following Questions:_

14. What is the variance estimate for the amount of variance due to between schools (group) differences? 

15. What is the variance estimate for the amount of variance due to within school (group) differences? 

16. Calculate the Intraclass correlation coefficient.  Interpret.

17. Now using the solution for fixed effects portion of the output, interpret the intercept.  

18.	Run the following code, which will allow you to assess whether verbal IQ, a level 1 variable, is associated with langpost, after accounting for the fact that individuals are nested within schools.
For easier interpretation we will center IQ verbal variable around the grand mean.

```{r}
sb$iq_center <- scale(sb$IQ_verb, scale = FALSE)
```

```{r}
fitmix2 <- nlme::lme(langPOST ~ iq_center, data = sb, random = ~1 |  schoolNR)
summary(fitmix2)
intervals(fitmix2)
```

Extract the variance and correlation components from an `lme` class object. 

```{r}
VarCorr(fitmix)
VarCorr(fitmix2)
```

19. Interpret the intercept.


20. Interpret the fixed effect estimate for the IQ verbal (centered) term.


21. Write out the equation for the model specified above.


22. How much of the within school variation in langpost is explained by IQ verbal?

