---
title: "Meta Analysis in R"
date: "March 8, 2016"
output: html_document
---

# Meta-analysis
Dataset source: Borenstein et al. (2009)

1. Compute precision for each study (1/variance)

2. Fixed Effects Model 
* compute weights: (1/variance)
* pooled effect: $M = \sum_{i=1}^k W_i * effect size_i/\sum_{i=1}^k W_i$
* its standard error SE: $\sqrt{1/\sum W_i}$
* 95% CI for the pooled effect size: ES $\pm$ 1.96 * SE

### Report pooled effect size and its 95% CI
* Compute: Q statistic as  $Q = \sum_{i=1}^k W_i (effects size_i-summary effect)^2$
    * df as the number of studies (k) - 1
    * _p_-value 
* Compute $I^2$ statistic (*100)

**Interpret $Q$ and $I^2$**

3. Random Effects
* compute weights $1/(d^2+\tau^2)$ to estimate $\tau^2$ we use the DerSimonian-Laird method 
$$
T^2 = (Q - df)/C
$$
where $C = \sum W_i-(\sum W_{i}^2/\sum W_i)$
* pooled effect $M = \sum_{i=1}^k W_i * effect size_i/\sum{i=1}^k W_i$
* 95% CI for the pooled effect size: ES $\pm$ 1.96 * SE
* z-value to test the null hypothesis that the mean effect is zero: Z = M/SE

**Report pooled effect size and its 95% CI**

# Continuous outcome

```{r, warning=FALSE}
library(readxl)
bor <- read_excel("contmeta.xlsx")
bor
```

Load packages

```{r, warning=FALSE}
suppressPackageStartupMessages(library(metafor))
library(compute.es)
library(MAd)
```

## Compute Standardized Mean Differences
```{r}
bor1 <- compute_dgs(n.treat, m.treat, sd.treat, n.plac, m.plac, sd.plac, data = bor, denom = "pooled.sd")
```

## Fixed Effects

Compute the weights fixed effects
```{r}
bor2 <- wgts(g, var.g, bor1)
```

Sum of Weights Fixed Effects `sumWi`
```{r}
cat("sumWi =", sumWi <- sum(bor2$wi))
```

Weights Fixed Effects * Effect Size `wfees`
```{r}
cat("sumWfees =", sumWfees <- sum(bor2$wiTi))
```

pooled effect size `pes`
```{r}
cat("pes =", pes <- sumWfees/sumWi)
```

standard error of pooled ES `sepes`
```{r}
cat("sepes =", sepes <- sqrt(1/sumWi))
```

95% CI `lower` and `upper` limit
```{r}
cat("lower 95% CI =", lower <- pes-1.96*sepes)
cat("upper 95% CI =", upper <- pes+1.96*sepes)
```

```{r}
rbind(pes, sepes, lower, upper)
```

## Q statistic

```{r}
bor3 <- cbind(bor2, espes = bor2$g-pes, espes2 = (bor2$g-pes)^2, espes2wi = ((bor2$g-pes)^2)*bor2$wi)
cat("qstat =", qstat <- sum(bor3$espes2wi))
cat("df =", df <- nrow(bor3)-1)
cat("p-value =", pchisq(qstat, df, lower.tail = FALSE))
```

Compute $I^2$

```{r}
cat("I^2 = ", istat <- (qstat-df)/qstat*100)
```

## Random Effects

```{r}
bor3$wi2 <- bor3$wi^2
sumWi2 <- sum(bor3$wi2)
cat("C =", bigC <- sumWi-sumWi2/sumWi)
cat("Tau^2 =", tau2 <- (qstat-df)/bigC)
```

Weights random effects `wi.tau` and Weights random effects * Effect Size (`d`) `wiTi.tau`

```{r}
bor3[, c("wi.tau", "wiTi.tau")]
```

Pooled effect size 2
```{r}
cat("pooled effect size =", pes2 <- with(bor3, sum(wiTi.tau)/sum(wi.tau)))
cat("standard error of pooled ES =", sepes2 <- with(bor3, sqrt(1/sum(wi.tau))))
cat("95% CI lower =", lower2 <- pes2-1.96*sepes2)
cat("95% CI upper =", upper2 <- pes2+1.96*sepes2)
cat("Z test for Ho M=0 is", z <- pes2/sepes2)
cat("p-value = ", 2*(1-pnorm(abs(z))))
```
