---
title: "Propensity Score Matching in R"
date: "March 22, 2016"
output: html_document
---

__Dataset:__ Chapt3 (Source: Analysis of Observational Health Care Data using SAS)

__Source:__ Analysis of Observational health care data using SAS

__Data Description:__ A total of 5162 patients undergoing invasive cardiac procedure in Ontario, Canada (data obtained from clinical registry maintained at the Cardiac Care Network of Ontario (CCN). Patients underwent a PCI with placement of either DES or BMS. Data was modified for this assignment.

|Variable name |Description|
|---------------|-----------|
|acutemi |Acute Myocardial Infraction (1=yes, 0=no) |
|cens | Not used in this analysis|
|diabetic | Diabetes (1-yes, 0=no)|
|female | Gender (0-male, 1-female)|
|height | height|
|mort6mo | Death within 6 months|
|stent | Stent (1-yes, 0=no)|
|trtm | treatment variable (DES=1 vs. BMS=0)|
|tte | Not used in this analysis|

Load the dataset using `haven`:

```{r}
library(haven)
library(dplyr)
chap3 <- read_sas("data/Chapter3_Example_Data.sas7bdat")
head(chap3)
dim(chap3)
sapply(chap3[, c(3,5,6,7)], table)
```

1. Assess the distribution of baseline characteristics:

```{r}
group_by(chap3, trtm) %>% summarize(avg_ht = mean(height), n(),
                                    sd = sd(height), min = min(height),
                                    max = max(height))
t.test(height ~ trtm, data = chap3)
lapply(chap3[, c("stent", "female", "diabetic", "acutemi")], function(var) {
    cbind(xtabs(~ trtm + var, data = chap3),
          prop.table(table(chap3$trtm, var)))
    })
```

__Note:__ You may also use the `gmodels` package to create tables.

```{r}
library(gmodels)
CrossTable(chap3$trtm, chap3$stent, prop.chisq = FALSE)
chisq.test(table(chap3$trtm, chap3$stent))
```

2. R code for fitting a propensity score model

```{r}
fit1 <- glm(trtm ~ height + stent + female + diabetic + acutemi,
            data = chap3, family = binomial)
summary(fit1)
```

Take the predicted probabilities of being treated along with treatment.

```{r}
ps_df <- cbind(data.frame(ps_score = predict(fit1, type = "response"),
               chap3))
head(ps_df)
ps_df$treatment <- factor(ps_df$trtm, levels = c(0,1),
                          labels = paste("Treatment type:", c("BMS", "DES")))
```

Using the `lattice` package:

```{r}
library(lattice)
histogram(~ps_score | treatment, data = ps_df,
          type = "count",
          col = "darkgreen",
          xlab = "Probability of receiving treatment")
```

Using the `ggplot2` package (optional but recommended):

```{r}
library(ggplot2)
ggplot(ps_df, aes(x = ps_score)) + geom_histogram(color = "white", binwidth = 0.02) + facet_wrap(~treatment) +
  xlab("Probability of receiving treatment") + theme_bw()
```

Fit a logistic regresssion using the propensity score and add the generalized logit of the score

Logistic regression for comparing risk of death within 6 months:

```{r}
library(gtools)
fit2 <- glm(mort6mo ~ trtm + ps_score + height + stent + female + diabetic + acutemi, data = ps_df, family = binomial)
summary(fit2)
ps_df$logit_ps <- logit(ps_df$ps_score)
```

Patients will now be matched on the logit of the propensity score using clipper of 0.2 standard deviations of the logit of the propensity score.

Compute the standard deviation of the logit of the propensity score: 
```{r}
sd(ps_df$logit_ps)
```

Run the `matchit` function with method set to `nearest` (considered greedy matching when `ratio` argument not defined - _default = 1_): 

__Note:__ `caliper` defined as the number of standard deviations of the distance measure within which to draw control units.

```{r}
suppressPackageStartupMessages(library(MatchIt))
m.out <- matchit(trtm ~ height + stent + female + diabetic + acutemi, caliper = 0.2, method = "nearest", data = chap3)
summary(m.out)
m.dat <- match.data(m.out)
```

The variable `distance` in the dataset indicates the propensity score: 

```{r}
head(m.dat)
```

