---
title: "Lab Exercise Chi Square"
author: "PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading SPSS Data

Use the `haven` package and the `read_spss` function to load SPSS data into the
R session:

```{r}
library(haven)
read_spss("stenosis_2014.sav")
```

1. Ensure that the data file is in the same folder as this RMarkdown file for
staightforward read-in when knitting.
2. You can find the full path to the file by using `file.choose()` and locating
the file in your folders.

```{r}
data("HairEyeColor")
head(HairEyeColor)
femaleHEC <- HairEyeColor[, , "Female"]
```

These are the expected values from the `chisq.test` function.

```{r}
res <- chisq.test(femaleHEC)
res$expected
```

## Manual calculation of expected values (advanced)

To calculate manually, you need to use `rowSums` and `colSums` at
each cell and then divide by the grand total:

$ e = \frac{row.sum \times col.sum}{grand.total} $

For the first cell:

```{r}
(rowSums(femaleHEC)[1] * colSums(femaleHEC)[1]) / sum(femaleHEC)
```

As you can see this matches with `res$expected[1]`: `r res$expected[1]`.

To avoid doing all these calculations manually we can get a little help from
R.

First we get all the combinations of `rowSums` and `columSums` pairs for each
cell:

```{r}
( allCombos <- expand.grid(rowSums(femaleHEC), colSums(femaleHEC)) )
```

Note that we are getting the combinations down the columns of the matrix
`femaleHEC`, (i.e., row1 & colsum1, row2 & colsum1, row3 & colsum1, row4 &
colsum1). This indicates that our `byrow` argument in the `matrix` function
below will be `FALSE` meaning the results are going down the columns and not
across the rows.

Now we loop through all of these values and apply the expected value equation
at the same time:

```{r}
expecteds <- apply(allCombos, 1, function(x) x[[1]] * x[[2]])  / sum(femaleHEC)
```

Now we can reconstruct the matrix of results with the `matrix` function
and some arguments:

```{r}
exp_mat <- matrix(expecteds, byrow = FALSE, nrow = 4,
    dimnames = list(rownames(femaleHEC), colnames(femaleHEC)))
round(exp_mat, 5)
```

Comparing to the original results:

```{r}
res$expected
```

Looks good!

## Percentages by rows

The data is already a 'tally' table so we should only run `prop.table` to get
percentages. Had we had a dataset, we would've had to run the `table` function.

Now let's see the row percentages:

```{r}
femaleHEC
prop.table(femaleHEC, margin = 1L)
```


# State you null and alternative hypothesis:

Chi-square test examines whether rows and columns of a contingency table are
statistically significantly associated.

* Null hypothesis (H0): the row and the column variables of the contingency
    table are independent.
* Alternative hypothesis (H1): row and column variables are dependent.

from http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r

Now we run the test:

```{r}
reschi <- chisq.test(femaleHEC)
reschi
reschi$p.value
```

Reject the $H_0$ hypothesis. Female hair and eye color are not independent.

# Relative Risk

Let's create an example dataset from the University of Virginia:

```{r}
MI <- matrix(c(189, 104, 10845, 10933), nrow = 2, dimnames =
list(Group = c("Placebo","Aspirin"), MI = c("Yes","No")))
```

Source: https://data.library.virginia.edu/comparing-proportions-with-relative-risk-and-odds-ratios/

Let's have a look at a mock (?) example of an RTC of Myocardial Infarction and
aspirin treatment (from Agresti, 1996):

```{r}
MI
prop.table(MI, margin = 1)
```

## Relative Risk calculation

```{r}
prop.out <- prop.table(MI, margin = 1)
# relative risk of placebo vs. aspirin
prop.out["Placebo", "Yes"] / prop.out["Aspirin","Yes"]
```

To get the 95% confidence interval, we use the `epitools` package:

```{r}
# run this line in the console if you don't have the package installed
# install.packages("epitools")
library(epitools)
rr.out <- riskratio(MI, rev = "b")
rr.out$measure
```

Hypothesis test for Relative Risk:

$H_0: RR = 1$
$H_a: RR \neq 1$

## Odds Ratio

### State the hypothesis for the test:

$H_0: OR = 1$
$H_a: OR \neq 1$

Use the `oddsratio` function in the `epicalc` package:

```{r}
oddsratio(MI)$measure
```

