---
title: "Exploring Data and Graphs with Numerical Summaries"
author: "PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

To get familiar with R Markdown, open the boiler plate introduction to 
R Markdown by going to `File` > `New File...` > `R Markdown`

## R Scripts vs R Markdown

R scripts are purely R code and R Markdown allows you to combine instructions
like these with R code. R scripts end in the `.R` file extension. R Markdown
files end in `.Rmd` file extension.

## Installing R packages

Try installing the `haven` package in the console below.
*Important*. Package installations should not be part of your code but run
interactively (in the console below).

```{r,eval=FALSE}
install.packages("haven")
```
## Loading a package

In R you load packages using: 

```{r}
library(haven)
```

Note that this command will not work if you have not installed the package
first.

## File paths

Finding your files on your computer is important so that you can tell R
where to go look for it. The easiest way to go about it is to put the file
in the same folder as this R Markdown file. The other way is to get the
file path from your operating system and tell R where the file is located.
In the example below, I used the easier option.

## Importing data

With the `haven` package, we can import SPSS data into our R session:

```{r}
birth <- read_spss("NCBIRTH800mod.sav")
```
# Lab 1 Data Exploration

Now that we have the data loaded, we can explore it. I recommend a set of
functions to explore the data. The `head` function gives you a small view
of the top of the dataset. 

```{r}
head(birth)
```

Use the `dim` function to see the dimensions (rows by columns) of the data:

```{r}
dim(birth)
```

There are 800 rows and 15 columns in the dataset.

# Data Types

R has 3 basic data types:

## Logical 

```{r}
TRUE
FALSE
```

## Numeric

```{r}
1
2
3
```

## Character

```{r}
"red"
"orange"
"yellow"
```

Notice that each line is a command. We will see how to combine these into one
line using the `c` (combine) function.

## Lab 2

In this example, what percentage of mothers smoked during pregnancy? Display
these data using a bar chart as well. 

First we create a 'table' variable from the smoke variable by running the
`table` function on the variable. We extract variables using the dollar sign
symbol `$` from the dataset. We then use `<-` to store the 'table' variable
into our R session. 
We then run `prop.table` on the result with some adjustments to make it into
a percent (multiply by 100 and round decimal places).

```{r}
## create a 
(smoke_table <- table(birth$smoke))
## proportions in percentages
round(prop.table(smoke_table), 3) * 100
```

Because our dataset did not contain any labels as imported, we will have to 
add them ourselves.

### Factors

In R, numeric variables with labels are what we call factors. To create a factor
we use the `factor` function. As you guessed it, we use the `factor` function
on the single _extracted_ variable. Remember we need to assign the output of
this function to a variable using the `<-` assignment operator. 

```{r}
smoked <- factor(
    x = birth$smoke,
    levels = 0:1,
    labels = c("Did not smoke", "Smoked")
)
```

As you can see, R does not care about extra spaces or carriage returns in our
code but we do because we want to be able to easily read the code.

Notice that this variable is not part of the dataset. It only lives within our
session as reflected by the Global Environment on the top right pane in
RStudio. If we want to include this in our dataset, we have to add it as
a variable:

```{r}
birth$smoked <- smoked
head(birth)
```

## Histograms

```{r}
barplot(smoke_table)
```

Notice that our original 'table' variable (`smoke_table`) did not have any
lables. We can now use are updated variable `smoked` to FIRST create a
table from it and then feed it to barplot:

```{r}
barplot(table(smoked))
```

## Crosstabs

In R, we are capable of creating crosstabs from two variables using
formula notation using the `xtabs` function. Let's see how this looks.

```{r}
xtabs(~ ethnmom + smoke, data = birth)
```

We may also use the familiar `table` function to do the same:

```{r}
table(birth$ethnmom, birth$smoke)
```

Notice that in both commands, the dataset has to be referenced somehow so that R
knows where the variables are coming from `data = birth` and `birth$smoke`.

### Proportions for crosstabs

Again we use the `prop.table` function to get the proportions but first,
let's create a table variable from one of these commands:

```{r}
eth_by_smoke <- xtabs(~ ethnmom + smoked, data = birth)
```

Now we use the `prop.table` command with the included argument `margin = 1`:
This tells R to create the proportions across the rows. We also round and 
multiply by 100 to get a percentage.

```{r}
round(prop.table(eth_by_smoke, margin = 1), 3) * 100
```

## Grouped Barplot

```{r}
barplot(eth_by_smoke, beside = TRUE)
```

# Numerical summaries

In this sample, what was the average birth weight (in grams)? Standard
Deviation, Minimum and Maximum? Display this data using a histogram as well.
Is the distribution of birth weight symmetric or skewed?

```{r}
summary(birth$tgrams)
```

We use a histogram to visualize numerical continuous data.

```{r}
hist(birth$tgrams)
```

## Getting a normal curve overlay

We run this code to get a normal distribution curve on top of the
histogram. We can also use other facilities that may make this easier in
packages such as `ggplot2`.

```{r}
x <- birth$tgrams
h <- hist(x, breaks=10, 
    xlab="Baby weight in grams", main="Histogram with Normal Curve")
xfit <- seq(min(x),max(x),length=40)
yfit <- dnorm(xfit,mean=mean(x),sd=sd(x))
yfit <- yfit*diff(h$mids[1:2])*length(x)
lines(xfit, yfit, col="red", lwd=2) 
```

## Grouped summaries

In this sample, what was the distribution of birth weight (in grams) by mother’s
smoking status?

```{r}
no_smoke <- birth$tgrams[birth$smoke == 0]
yes_smoke <- birth$tgrams[birth$smoke == 1]
summary(no_smoke)
summary(yes_smoke)
```

### Grouped boxplots

Here we use formula notation to indicate to R how to take numerical data
and what variable to use when splitting the data.

On the left hand side of the tilde `~` we enter the name of the numerical 
variable we will be using to compare distributions across groups. The
`smoked` variable will be the grouping variable and again we tell R that
the data comes from the birth dataset by doing `data = birth`.

```{r}
boxplot(tgrams ~ smoked, data = birth)
```

# Questions (answers submitted in Quiz 2):

1. Describe the basic demographic characteristic of the sample

* Mother’s age [variable: `mage`]
* Marital status [variable: `marital`]
* Race of mother [variable: `ethnmom`]

2. What was the median mother’s age for mothers who smoked during pregnancy vs.
mothers who did not smoke during pregnancy?

* Variables: `mage` and `smoked`

3. Were there any outliers in mother’s age in either of the smoking status
groups (smokers vs. non-smokers) [Hint: inspect the side-by-side box plot].
