---
title: "Logistic Regression"
author: "PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# North Carolina Birth Dataset

Recall that the North Carolina State Center for Health Statistics makes
publicly available birth and infant death data for all children born in the
state of North Carolina. This comprehensive dataset for the births in 2001
contains 120,300 records. The data in "NCBIRTH800mod.sav" which we have used
previously represents a random sample of 800 of those births and selected
variables.

# Research Question 1

Is low birth weight associated with mother’s smoking status during pregnancy
after accounting for gestational age?

# Loading the data

```{r}
library(haven)
ncbirth <- read_spss("~/Downloads/NCBIRTH800mod.sav")
```

# Contingency Tables

There are many ways to create crosstabs in R. One way is to use the `xtabs`
function with formula notation. A formula is a convenient way of telling
R how variables are related in a model.

## Formula notation

In R, the tilde `~` is what defines a formula. It is essentially a stand-in
for the function that uses formulas to know how to use the associated
variables.

```{r}
Y ~ X1 + X2
```

In the above formula, Y, X1, and X2 should be variables in a dataset. The
dataset is usually identified by a `data =` argument

### Example

For crosstabs, we are using two X variables (if you will) to create a
cross-tabulation (e.g., `~ X1 + X2`).

```{r}
xtabs(~ smoke + low, data = ncbirth)
```

## table

You can also use the `table` function with two categorical variables / vectors.

```{r}
table(ncbirth$smoke, ncbirth$low)
```

## Row percentages

Here we use the `prop.table` with the `margin = 1L` argument to get row
proportions:

```{r}
tab1 <- xtabs(~ smoke + low, data = ncbirth)
prop.table(tab1 , margin = 1L)
```

You may also want to round:

```{r}
round( prop.table(tab1 , margin = 1L), 2)
```

### Summary statistics

% low birth weight infants for mothers who smoked during pregnancy: 13.2%
% low birth weight infants for mothers who did not smoke during pregnancy: 8.0%

## Chi-Square test of independence

```{r}
prop.test(tab1, correct = FALSE)
```

## Odds Ratio

You can calculate this manually or using the examples we have seen
previously:

```{r}
a <- tab1[1, 1]
b <- tab1[1, 2]
c <- tab1[2, 1]
d <- tab1[2, 2]
(a*d) / (b*c)
```

With the `epitools` package:

```{r}
library(epitools)
oddsratio(tab1)$measure
```

### Interpretation

The odds of low birth weight infants are 1.73 times higher among mothers who
smoked during pregnancy as compared to those who did not smoke. This difference
is not significant (p = 0.074).

# Logistic Regression

We use the `glm` function to run a logistic regression in R. Here our knowledge
of formulas becomes useful:

```{r}
fit <- glm(low ~ smoke + weeks, data = ncbirth, family = "binomial")
summary(fit)
```

We can create our summary table manually:

```{r}
ctable <- coefficients(summary(fit))
coeffs <- exp(coef(fit))
ints <- exp(confint(fit))
round( cbind(ctable, ORs = coeffs, ints), 3)
```

Interpretation: The odds of low birth weight babies are 1.96 times higher among
mothers who smoked during pregnancy as compared to those who did not smoke,
after adjusting for gestational age.

(CI for OR (.93, 4.11)), this association is not significant (p=.075).

# Research Question 2

Is low birth weight associated with mother’s ethnicity, after accounting for
gestational age, mother’s age and smoking status during pregnancy?

## factors

When using the `haven` package to read in data, creating a factor from the data
is quite easy:

```{r}
ethnmom_f <- as_factor(ncbirth$ethnmom)
table(ethnmom_f)
```

To show you how to do this without a `haven` loaded dataset, we use the
factor function. First we have a look at the levels using a table:

```{r}
table(ncbirth$ethnmom)
```

Then we use this information to recode. Usually, we get the value labels
information from the study's codebook. We enter the labels as the input to
the `labels =` argument as a character vector:

```{r}
ethnmom_f <- factor(ncbirth$ethnmom, levels = 1:4,
    labels =  c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Other")
)
table(ethnmom_f)
```

Either way, we have a factor that R can use to run our model.

## relevel

To check the reference category in our factor, we use the `contrasts` function:

```{r}
contrasts(ethnmom_f)
```

To change the reference category in our factor, we use the `relevel` function:

```{r}
ethnmom_f <- relevel(ethnmom_f, ref = "Non-Hispanic White")
```

Note that our factor variable exists independent of the data. To insert the
variable into our dataset, we use this notation:

```{r}
ncbirth$ethnmom_f <- ethnmom_f
```

On the right, we have our factor variable and on the left we are adding
it to our existing dataset `ncbirth`.

# Bivariable regression (unadjusted)

```{r}
fit2 <- glm(low ~ ethnmom_f, data = ncbirth, family = "binomial")
summary(fit2)
cbind(OR = exp(coef(fit2)), exp(confint(fit2)))
```

In the unadjusted (crude) analysis:
* The odds of having low birth babies were higher for Non-Hispanic Black
mothers, as compared to Non-Hispanic White mothers (reference category):
OR=2.01, 95%CI (1.47, 3.45), p=0.11
* There were no statistically significant differences in the probability of
having low birth weight babies between Hispanic vs. Non-Hispanic White mothers
(reference categories): OR=.92, 95%CI(.39, 2.32), p=.916.

# Multivariable Regression (adjusted)

You can breakdown the code below as in the first example or run it all
in one go. Essentially, we are taking the summary table (from `summary(fit3)`)
and adding Odds Ratio and confidence intervals columns.

```{r, warning=FALSE}
fit3 <- glm(low ~ ethnmom_f + smoke + weeks + mage,
    data = ncbirth, family = "binomial")
summary(fit3)
round(cbind(
    coefficients(summary(fit3)), OR = exp(coef(fit3)), exp(confint(fit3))
), 3)
```

## Interpretation

There are no statistically significant differences in the probability of having
low birth weight babies between Non-Hispanic Black mothers vs. Non-Hispanic
Black mothers (p=.210) and Hispanic vs. Non-Hispanic White mothers (p=.664),
after adjusting for gestational age, mother’s age and smoking during pregnancy.

## Note

The confidence intervals in R do not exactly match those in SPSS but the
results and inferences are largely the same.

