---
title: "Multiple Linear Regression"
author: "PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example dataset `mtcars`

For more information about the dataset, type in `?mtcars` in the console.

```{r}
data(mtcars)
# ?mtcars
```

## Correlation Test

To see how two continuous variables relate to each other, we use a Pearson
correlation:

```{r}
cor.test(mtcars$mpg, mtcars$wt)
```

## Simple Linear Regression

We want to answer our research question. Is a car's weight associated
with fuel efficiency (as measured by miles per gallon)?

```{r}
fit1 <- lm(mpg ~ wt, data = mtcars)
summary(fit1)
```

# Assumptions check

## Checking for normality in the outcome variable

```{r}
hist(mtcars$mpg, breaks = 10)
```

## Quantile-Quantile Plot

We also can check for normality using a Q-Q plot:

```{r}
qqnorm(mtcars$mpg, pch = 1)
qqline(mtcars$mpg)
```

The points should fall on the line to be considered normally distributed.

## Homoscedasticity

The assumption is that random variables have the same finite variance. That
means that this plot should not have a specific pattern to it. The spread
of the data points is approximately constant.

```{r}
plot(fitted(fit1), residuals(fit1))
```

## Normality of residuals

The residuals should have a mean of zero and be approximately normal:

```{r}
hist(residuals(fit1), breaks = 10)
```

Checking the mean:

```{r}
mean(residuals(fit1))
```

It is a very small number (practically zero).

## Independence

Is the data independent? Probably not, the cars in the dataset are not all from
different manufacturers. Some cars may be highly correlated because of this. 

# Exploratory Data Analysis

To visually compare all variables against each other, we can use this 
`pair.panels` matrix. To see the relationship between any two variables find
the intersecting square based on the row and the column variables, e.g.,
the correlation for `mpg` and `hp` can be found in the box in row 1 column 4
(correl coef: `-0.78`).

```{r}
library(psych)
pairs.panels(mtcars, cex = 1.5)
```

# Multiple Linear Regression

To include multiple variables in the regression, modify the formula inside
the `lm` function:

```{r}
fit2 <- lm(mpg ~ wt + hp, data = mtcars)
summary(fit2)
```

You can run the same steps as above with a multiple linear regression to check
the assumptions.

We will now go over a few additional notes for working with the SPSS
dataset and factors in a linear regression model.

# SideNotes

## Loading an SPSS dataset into R

I recommend using the `haven` package and the `read_spss` function in that
package:

```{r}
library(haven)
ncbirth <- read_spss("NCBIRTH800.sav")
```

Note. Make sure that the data file is in the same folder as the RMarkdown file
for easier knitting. To learn how to knit, check out the RStudio RMarkdown
tutorials.

### including categorical variables as factors

To include categorical variables, use the factor function to tell R
which variables are categorical. Here we tell R that the number of cylinders
(`cyl`) is categorical:

```{r}
fit3 <- lm(mpg ~ wt + hp + factor(cyl), data = mtcars)
summary(fit3)
```

You should get a coefficient for each category that is not the reference.
In R, the lowest value is taken as the reference. One way to check is to 
use the `contrasts` function on a `factor` variable.

```{r}
table(mtcars$cyl)
cyl_f <- factor(mtcars$cyl)
contrasts(cyl_f)
```

Rows with all zeros across them are the reference, in this case `4` cylinders
is the reference. 
To change that, we use the `relevel` function and check again with `contrasts`.

```{r}
cyl_f <- relevel(cyl_f, ref = "8")
contrasts(cyl_f)
```

## Data cleaning factors

Because we used the `haven` package, the data retains the value labels for
categorical variables. We can use this  information to clean up and create
our factor variables. First we check the variable categories with a `table`
function:

```{r}
table(ncbirth$ethnmom)
```

We then use the attributes function to extract the labels from the variable.
Note that this only works on SPSS datasets as read in with the `haven` package.

```{r}
attributes(ncbirth$ethnmom)$labels
```

We can assign this to a variable for later use:

```{r}
ethnmom_levels <- attributes(ncbirth$ethnmom)$labels
```

Note that this variable is a numeric vector with names:

```{r}
class(ethnmom_levels)
names(ethnmom_levels)
```

Now we create our factor variable with this information:

```{r}
ethnmom_f <- factor(ncbirth$ethnmom,
    levels = ethnmom_levels, labels = names(ethnmom_levels))
```

As always, we check our work:

```{r}
table(ethnmom_f)
```

Now we can add this variable back into the data. We can use the same name as
the variable that we created. On the left hand side we have the variable that
we are creating in the data and on the right we have the actual factor variable
that we created above.

```{r}
ncbirth$ethnmom_f <- ethnmom_f
```

Checking our work...

```{r}
head(ncbirth)
```

Now we see the variable we added to the extreme right of the dataset!
