---
title: "One-sample tests"
author: "PUBH 614"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Open NCBIRTH800mod dataset

Recall that the North Carolina State Center for Health Statistics makes
publicly available birth and infant death data for all children born in the
state of North Carolina. This comprehensive dataset for the births in 2001
contains 120,300 records. The data in `NCBIRTH800.sav` represents a random
sample of 800 of those births and selected variables.

```{r}
library(haven)
ncbirth <- read_spss("~/Downloads/NCBIRTH800mod.sav")
```

In 2001, smoking prevalence among women in North Carolina was 23.1% (Tobacco
Prevention and Control Branch & State Center for Health Statistics, 2002,
p.2)^[Tobacco Prevention and Control Branch & State Center for Health
Statistics. (2002). BRFSS News Brief - North Carolina. Who’s Still Smoking in
NC? Retrieved from
http://www.tobaccopreventionandcontrol.ncdhhs.gov/data/SurveillanceDataFiles/ArchiveFactSheets/adult/AdultShortReports/brfssgaso02.pdf].

Conduct a two sided hypothesis test ($\alpha$ = .05) to determine if smoking
prevalence among pregnant women is different than that of the general female
population in North Carolina.

## State the null and alternative hypotheses:

$H_0: p = 0.231$
$H_a: p \neq 0.231$

Would it be appropriate to conduct a "large sample" z-test?

What is the sample size?

This shows you the dimensions of the data. The first number corresponds to
the rows and the second to the
columns. There are 800 observations
and 15 variables in the data.

```{r}
dim(ncbirth)
```

Create the frequency table and proportions table with `table` and `prop.table`
respectively.

```{r}
table(ncbirth$smoke, useNA = "always")
prop.table( table(ncbirth$smokef) )
```

Note that there are two values that are missing `NA` in our data.

Since n*p (`798 * .231`) and n*(1-p) (`798 * .769`) are both greater than 5, it
is ok to use the large-sample z-test to test the hypotheses of interest.

Based on the documentation of `?binom.test`, we want to have a tally of
successes and failures in that order. To get a tally we use the `table`
function and reverse the order with `rev` to arrange successes first.

```{r}
smokes <- rev(table(ncbirth$smoke))
zres <- prop.test(x = smokes, p = 0.231, alternative = "two.sided", correct = FALSE)
zres
```

Now we confirm that our probability of successes `p` matches the mean below:

```{r}
## mean should match prob. of success
mean(ncbirth$smoke, na.rm = TRUE)
```

To get the Z-test statistic, you take the square root of the `X-squared`
value in the output:

```{r}
sqrt(zres$statistic)
```

We know that our sample proportion, `0.142`, is less than the population
estimate `0.231` (smoking prevalence among women in NC) then we can conclude
that the Z-test statistic is *negative*, i.e., `r -sqrt(zres$statistic)`.

Remember to set the value of the `alternative` argument in the `prop.test`
function based on whether you are conducting a one-sided or two-sided test.


## Conclusion

Conclusion: Since p-value associated with this test is _p_ < .001 (which is
smaller than $\alpha$ = 0.05), we reject the null hypothesis `Ho: p = .231` in
favor of the alternative hypothesis `Ha: p ≠ .231`. We conclude that the
proportion of women in North Carolina who smoke while pregnant is not `.231`; in
fact, we found that it is significantly less (95% CI: .12, .17) (for CI see
`zres$conf.int`).

To calculate the standard error, we use the formula:

$\sigma_{\bar{x}} = \sqrt{\frac{\sigma^2}{n}}$

```{r}
sqrt(sd(ncbirth$smoke, na.rm = TRUE)^2 / 798)
```

# One-sample test for means

Assume the average birth weight for a full-term newborn in the United States is
120 ounces. Conduct a two sided hypothesis test ($\alpha$ = 0.05) to determine
if the average birth weight of North Carolina newborns is significantly
different than that of the United States.

## State you null and alternative hypothesis: 

$H_0: \mu = 120$ ounces
$H_a: \mu \neq 120$ ounces

Plot the distribution of birth weights in the sample:

```{r}
hist(ncbirth$tounces, breaks = 30, main = "Weight of the child (ounces)")
```

Would it be appropriate to conduct a t test to test the hypotheses above? We
should be worried about the skew in the data. We will come back to this issue in
a moment. For the purpose of illustration we will for now consider all newborns,
i.e. full-term infants and premature infants, variable `tounces`).


First some descriptives:

```{r}
summary(ncbirth$tounces)
sd(ncbirth$tounces)
## calculate std. error manually
(stderr <- with(ncbirth, sd(tounces)/sqrt(length(tounces))))
```

We run the one-sample test with the `mu` argument set to our population
estimate.

```{r}
(onet <- t.test(x = ncbirth$tounces, mu = 120))
```

We can check the t-value manually as well:

```{r}
( mean(ncbirth$tounces) - 120 )/ stderr
```

Conclusion: Since p-value associated with this test is p<.001 (which is smaller
than $\alpha$ = 0.05), we reject the null hypothesis $H_0: \mu = 120$ ounces in
favor of the alternative hypothesis $H_a: \mu \neq 120$ ounces. We conclude that
the mean birth weight for newborns in North Carolina is not 120 ounces; in fact,
we found that it is most likely less -the 95% CI for the mean birth weight for
newborns in North Carolina is
(`r paste(round(onet$conf.int, 2), collapse = ", ")`) [see below] - given
repeated samplings, the true mean birth weight for newborns in North Carolina
will fall between `r paste(round(onet$conf.int, 2), collapse = ", ")` ounces 95%
of the time. Note the this CI does not include the hypothesized value of 120 oz,
indicating that the 120oz is NOT a plausible value for the mean birth weight for
newborns in North Carolina (hypothesis testing results and CI results agree).

```{r}
onet$conf.int
```

Now we will conduct the analysis on a subsample: the full-term newborns only. We
will select newborns using variable "premie" [1=infant was premature, 0= infant
was not premature].

```{r}
table(ncbirth$premie)
prop.table( table(ncbirth$premie) )
```

88.4% (n=707) of infants were full-term (not premature)

Now we take a the subset:

```{r}
fullterm <- ncbirth[ncbirth$premie == 0, ]
dim(fullterm)
```

State you null and alternative hypothesis: 

$H_0: \mu = 120$ ounces
$H_a: \mu \neq 120$ ounces

Sample size: n=707.

Plot the distribution of birth weights among the full-term infants:

```{r}
hist(fullterm$tounces, breaks = 30, main = "Weight of child (ounces)")
```

Descriptive stats:

```{r}
mean(fullterm$tounces)
sd(fullterm$tounces)
## calculate std. error manually
(stderr <- with(fullterm, sd(tounces)/sqrt(length(tounces))))
```

Now we run the one sample t-test again:

```{r}
(onet2 <- t.test(x = fullterm$tounces, mu = 120))
```

We can check the t-value manually as well:

```{r}
( mean(fullterm$tounces) - 120 )/ stderr
```

Conclusion: Since p-value associated with this test is p=.780 (which is greater
than $\alpha$ = 0.05), we fail to reject the null hypothesis $H_0: \mu = 120$
ounces. We did not find evidence that the mean birth weight for full-term
newborns in North Carolina is different from the mean birth weight for full-term
newborns in the U.S.

```{r}
onet2$conf.int
```

The 95% CI for the mean birth weight for full-term newborns in North Carolina is
(`r paste(round(onet2$conf.int, 2), collapse = ", ")`). Note the this CI does
include the hypothesized value of 120 oz, indicating that the 120 oz is a
plausible value for the mean birth weight for full-term newborns in North
Carolina (again hypothesis testing results and CI results agree).

# Questions:

Test the following hypotheses at $\αlpha$ = 0.05:

* The percentage of premature infants in North Carolina in 2001 was significantly lower than the percentage premature babies in the U.S. in 2001, which is reported to be 11.9% (http://www.cdc.gov/nchs/data/nvsr/nvsr51/nvsr51_02.pdf). [variable `premie`]
Make sure to sort "premie" variable in descending order.
* The mean mother’s age in North Carolina in 2001 was significantly different
than the mean mother’s age in the U.S. in 2000, which was reported as 24.9 years
old. (http://www.cdc.gov/nchs/data/nvsr/nvsr51/nvsr51_01.pdf). [variable mage] 

First, write your hypotheses and then use R to test them. 
