---
title: "TwoSampleTests"
author: "PUBH 614"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Open NCBIRTH800mod dataset

Recall that the North Carolina State Center for Health Statistics makes
publicly available birth and infant death data for all children born in the
state of North Carolina. This comprehensive dataset for the births in 2001
contains 120,300 records. The data in `NCBIRTH800.sav` represents a random
sample of 800 of those births and selected variables.

```{r}
library(haven)
ncbirth <- read_spss("~/Downloads/NCBIRTH800mod.sav")
# ncbirth <- haven::read_spss("~/data/NCBIRTH800mod.sav")
```

## Labels setup

The data from SPSS comes with labels. Let's use it to create something
that R can recognize.

```{r}
smoke_levels <- attributes(ncbirth$smoke)$labels
ncbirth$smoke <- factor(ncbirth$smoke, levels = smoke_levels,
    labels = names(smoke_levels))
```

Now we get a nicer output!

```{r}
table(ncbirth$smoke)
```

We'll do the same for low birthweight (`low` variable):

```{r}
low_levels <- attributes(ncbirth$low)$labels
ncbirth$low <- factor(ncbirth$low, levels = low_levels,
    labels = names(low_levels))
table(ncbirth$low)
```

# Two sample test for proportions

Are mothers who smoke during pregnancy more likely to have low birth weight
(<2500 grams) infants compared to mothers who do not smoke during pregnancy?

## State you null and alternative hypothesis: 
      
$H_0:   p1 – p2 = 0$ or $(p1 = p2)$
$H_a:  p1 – p2 \neq 0$ or $(p1 \neq p2)$

Would it be appropriate to conduct a "large sample" z test? 

# Test this hypothesis using R

First we create a table object using the `table` function with our exposure
and outcome of interest:

```{r}
(smlow <- table(ncbirth$smoke, ncbirth$low))
```

We then use `prop.table` with the `margin = 1` argument to get row percentages:

```{r}
prop.table( smlow , margin = 1)
```

What is the sample size? 

There were n=798 valid observations (2 observations were missing).

## Sample statistics

$n_1$ = 114 – # of mothers who smoked during pregnancy 
$n_2$ = 684 – # of mothers who did not smoked during pregnancy 

$\hat{p_1}$ = .13   [15/114 = .13]  [proportion of low birthweight babies for
smokers]
$\hat{p_2}$ = .08   [55/684 = .08]  [proportion of low birthweight babies for
non-smokers]

$\hat{p_1} - \hat{p_2}$ = .13- .08 = .05  [Point estimate of the difference
(p1-p2), also called the risk difference]

Since ($n_1p_1$ ≥ 5 & $n_1(1-p_1)$ ≥ 5) -> 114*.13=14.82≥ 5 & 114*.87=99.18≥ 5
and ($n_2p_2$ ≥ 5 & $n_2(1-p_2)$ ≥ 5) -> 684*.08=54.72 ≥ 5 & 684*.92=629.28≥ 5;
it is ok to use the large-sample z-test to test the hypotheses of interest.

Finally, we run a chi-square test of independence:

_Note_. We need to set `correct = FALSE` to avoid using Yates' Continuity
Correction.

```{r}
(tabres <- chisq.test(smlow, correct = FALSE))
```

R does not provide the value of the z-test directly – it can be obtained by
taking the square root of the Pearson Chi-square value: $\sqrt{3.197}$ = 1.79.

```{r}
sqrt(tabres$statistic)
```

Two-sided p-value is listed under "p-value". Recall that if the alternative
hypothesis was 1-sided (p1 – p2 >0   or p1 – p2 < 0), we would divide the
2-sided p-value by 2. Example here rounded to three decimal places:

```{r}
round(tabres$p.value / 2, 3)
```

Conclusion: Since p-value associated with this test is p=.074 (which is larger
than $\alpha$ = .05), we fail to reject the null hypothesis Ho: p1 – p2 = 0 in
favor of the alternative hypothesis $H_a: p1 – p2 \neq 0$. Notice that the
p-value is very close to the cut-off value of $\alpha$ = .05 – such a finding is
often reported as a trend towards a significant difference (in the expected
direction). This study should be replicated to clarify the relationship between
smoking during pregnancy and low birth weight.

## 95% CI in R

```{r}
prop.test(smlow, correct = FALSE)$conf.int
```

_Note_. Zero is within our confidence interval, indicating that (p1-p2) may
equal 0.

### Calculation of the Z-test

```{r}
zres <- prop.test(smlow, correct = FALSE)
## Z value
sqrt(zres$statistic)

## p-value
zres$p.value
```

# Two sample test for means (independent samples)

Do infants born to mothers who smoke during pregnancy have the same mean birth weight as those born to mothers who do not smoke?

## State the null and alternative hypothesis: 

$H_0:  \mu_1 - \mu_2 = 0$   ($\mu_1 = \mu_2$)
$H_a: \mu_1 - \mu_2 \neq 0$   ($\mu_1 \neq \mu_2$)

## Plot the distribution of birth weights by smoking status

```{r}
boxplot(tounces ~ smoke, data = ncbirth)
```

Multiple outliers show infants with very low birth weights, these are probably
premature infants.  Note that the majority of these infants are born to mothers
who did not smoke during pregnancy

First, we will conduct the analysis on the full sample i.e. full-term infants
and premature infants.

### Means by group

The groups are: 

```{r}
table(ncbirth$smoke)
```

There are many ways to get grouped summaries in R. `tapply` is one strategy:

```{r}
tapply(X = ncbirth$tounces, INDEX = ncbirth$smoke, FUN = mean)
tapply(X = ncbirth$tounces, INDEX = ncbirth$smoke, FUN = sd)
## manual calculation of the standard error
tapply(X = ncbirth$tounces, INDEX = ncbirth$smoke, FUN = function(x) {
    round(sd(x)/sqrt(length(x)), 3)
})
```

## t-test Results

```{r}
(tres <- t.test(tounces ~ smoke, data = ncbirth))
```

```{r}
tres$conf.int
```

### Sample statistics:
$n_1$ = 684 – # of mothers who did not smoked during pregnancy 
$n_2$ = 114 – # of mothers who did smoked during pregnancy 

$\bar{x_1}$ = 117.56 [sample mean birthweight for babies born to non-smokers]
$\bar{x_2}$ = 109.60   [sample mean birthweight for babies born to smokers]

$\bar{x_1} - \bar{x_2}$ = 117.56 - 109.60 = 7.96  [Point estimate of the
difference ($\mu_2 - \mu_2$)]

t-test=3.83, df=164.90, p-value<.001 

## Conclusion

Since p-value associated with this test is p<.001 (which is smaller
than $\alpha$ = 0.05), we reject the null hypothesis $H_0: \mu_1 - \mu_2 = 0$
in favor of the alternative hypothesis $H_a: \mu_1 - \mu_2 \neq 0$.

95% CI for $\mu_1 - \mu_2$ is (3.86, 12.06). With many resamplings of the
population the  difference in mean weight between babies born to mothers who
smoked during pregnancy and babies born to mothers who did not smoke during
pregnancy will fall  between 3.86 and 12.06 ounces 95% of the time. Note that
this CI does not include the hypothesized value of 0 for the mean difference,
indicating that the 0 is NOT a plausible value for the mean difference of
interest (hypothesis testing results and CI results agree).

# Subsample analysis

Now we will conduct the same analysis on a subsample: the full-term newborns
only. We will select full-term newborns using variable "premie" [1=infant was
premature, 0= infant was not premature].

First is the setup:

```{r}
premie_levels <- attributes(ncbirth$premie)$labels
ncbirth$premie <- factor(ncbirth$premie,
    levels = premie_levels, labels = names(premie_levels))
table(ncbirth$premie)
```

Note that now we have to match based on the label (i.e.,
`"Infant was not premature"`):

```{r}
fullterm <- ncbirth[ncbirth$premie == "Infant was not premature", ]
dim(fullterm)
```

## State the null and alternative hypothesis: 

$H_0:  \mu_1 - \mu_2 = 0$   ($\mu_1 = \mu_2$)
$H_a: \mu_1 - \mu_2 \neq 0$   ($\mu_1 \neq \mu_2$)

## Plot the distribution of birth weights by smoking status among full-term infants

```{r}
boxplot(tounces ~ smoke, data = fullterm)
```

### Means by group

The groups are: 

```{r}
table(fullterm$smoke)
```

There are many ways to get grouped summaries in R. `tapply` is one strategy:

```{r}
tapply(X = fullterm$tounces, INDEX = fullterm$smoke, FUN = mean)
tapply(X = fullterm$tounces, INDEX = fullterm$smoke, FUN = sd)
## manual calculation of the standard error
tapply(X = fullterm$tounces, INDEX = fullterm$smoke, FUN = function(x) {
    round(sd(x)/sqrt(length(x)), 3)
})
```

## t-test Results

```{r}
(tres <- t.test(tounces ~ smoke, data = fullterm))
```

```{r}
tres$conf.int
```

### Sample statistics:
$n_1$ = 606 – # of mothers who did not smoke during pregnancy 
$n_2$ = 99 – # of mothers who did smoke during pregnancy 

$\bar{x_1}$ = 121.49 [sample mean birthweight for babies born to non-smokers]
$\bar{x_2}$ = 112.69   [sample mean birthweight for babies born to smokers]

$\bar{x_1} - \bar{x_2}$ = 121.49 – 112.69 = 8.80  [Point estimate of the
difference ($\mu_2 - \mu_2$) among full-term babies]

t-test = 4.79, df = 135.29, p-value < .001  

## Conclusion

Since p-value associated with this test is p < .001 (which is smaller than
$\alpha$ = 0.05), we reject the null hypothesis $H_0: \mu_1 - \mu_2 = 0$
in favor of the alternative hypothesis $H_a: \mu_1 - \mu_2 \neq 0$.

95% CI for $\mu_1 - \mu_2$ is (5.17, 12.43). With many resamplings, the
difference in mean weight between babies born to mothers who smoked during
pregnancy and babies born to mothers who did not smoke during pregnancy wil fall
between 3.86 and 12.06 ounces 95% of the time. We arrived at the same
conclusions - on *average*, babies born to mothers who smoke tend to have lower
birth weight than babies born to mothers who do not smoke.


# Questions

Test the following hypotheses at $\alpha$ = 0.05:
* The average mother’s age is significantly different between mothers who smoked
during pregnancy and mothers who did not smoke during pregnancy.  [variables
`mage` and `smoke`]
* The percentage of premature infants born to mothers who smoked during
pregnancy is different than the percentage of premature infants born to mothers
who did not smoke during pregnancy.  [variables `premie` and `smoke`]

First Write your hypotheses and then use R to test them. 

