---
title: "Propensity Scores"
author: "BIOS 620"
rcte: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup load HELP RCT data

```{r}
library(haven)
rct <- read_spss("data/data_HELP.sav")
```

## Summary statistics

```{r}
summary(rct$cesd)
summary(rct$cesd[rct$homeless==0])
summary(rct$cesd[rct$homeless==1])
sd(rct$cesd)
sd(rct$cesd[rct$homeless==0])
sd(rct$cesd[rct$homeless==1])
```

## Histograms

```{r}
par(mfrow=c(3,1))
hist(rct$cesd, las=1)
hist(rct$cesd[rct$homeless==0], las=1)
hist(rct$cesd[rct$homeless==1], las=1)
```


## Density plots

```{r}
par(mfrow=c(3,1))
plot(density(rct$cesd), las=1)
plot(density(rct$cesd[rct$homeless==0]), las=1)
plot(density(rct$cesd[rct$homeless==1]), las=1)
```

### Boxplots

```{r}
par(mfrow=c(1,2))
boxplot(rct$cesd, ylab="Depression", las=1)
boxplot(cesd ~ homeless, data = rct, xlab="homeless", ylab="Depression", las=1)
```


# Describing the covariates (potential confounders): age

```{r}
t.test(rct$age, rct$homeless)
```

# Describing the covariates (potential confounders): female

```{r}
chisq.test(rct$female, rct$homeless)
```


# Describing the covariates (potential confounders): racegrp
```{r}
chisq.test(rct$racegrp, rct$homeless)
```


# Confouding model based causal inference (univariate analysis, unadjusted effect)

```{r}
fit <- lm(cesd ~ 1 + homeless,data=rct)
summary(fit)
```

# Confouding model based causal inference (multivariate analysis, adjusted effect)

```{r}
fit <- lm(cesd ~ 1 + homeless + age + female + factor(racegrp),data=rct)
summary(fit)
```

# Produce the propensity score

```{r}
m_ps <- glm(homeless ~ 1 + age + female + factor(racegrp),
    family=binomial, data=rct)
summary(m_ps)
```

Below, this propensity score is calculated by using `predict()` and create a
`data.frame` that has the propensity score as well as individual's actual
homeless status.

```{r}
prs_df <- data.frame(pr_score = predict(m_ps, type ="response"),
    homeless = m_ps$model$homeless)
head(prs_df)
```

After estimating the propensity score, it is useful to plot histograms of the
estimated propensity scores by homeless status: 

### Examining the region of common support of the scores

```{r}
library(ggplot2)
library(dplyr)
prs_df$homelessf <- factor(prs_df$homeless, levels = 0:1,
    labels = paste("Homeless status:", c("no", "yes"))
)
prs_df %>% ggplot(aes(x = pr_score)) +
  geom_histogram( color = "white") +
  facet_wrap(~ homelessf) +
  xlab("Probability of being homeless") +
  theme_bw()
```

Using 'base' plotting:

```{r}
par(mfrow = c(1, 2))
histogram(prs_df[prs_df$homeless == 0, "pr_score"])
histogram(prs_df[prs_df$homeless == 1, "pr_score"])
```


The method we use below is to find pairs of observations that have very similar
propensity scores, based on certain measures of **similarity**.

We use the package MatchIt for this. This package estimates the propensity score
in the background and then matches observations based on the method of choice.

```{r}
library(MatchIt)
rct_nomiss <- rct %>%  # MatchIt does not allow missing values
    dplyr::select(cesd, homeless, age, female, racegrp) %>%
    na.omit()

## ALTERNATIVELY
vars <- c("cesd", "homeless", "age", "female", "racegrp")
rct_nomiss2 <- rct[complete.cases(rct[, vars]), vars]
# identical(rct_nomiss, rct_nomiss2)

mod_match <- matchit(homeless ~ age + female + racegrp, method = "nearest", 
    replace=FALSE, M=1, data=rct_nomiss)
```

We can get some information about how successful the matching was using:

```{r}
par(mfrow = c(1, 1))
summary(mod_match) 
plot(mod_match) 
```

# Examine the matched data 

To create a data.frame containing only the matched observations, use the
`match.data()` function:

```{r}
dta_m <- match.data(mod_match) 
dim(dta_m)
dim(rct)
head(dta_m, 4)
```

# Visual inspection (e.g., with covariate Age)

```{r}
par(mfrow=c(1,1))

plot(dta_m$distance[dta_m$homeless==0],  dta_m$age[dta_m$homeless==0], 
    xlab="Propensity Score", ylab="Age", las=1)

points(dta_m$distance[dta_m$homeless==1],  dta_m$age[dta_m$homeless==1], pch=2)

legend("topleft", pch=c(1,2), c("Homeless 0", "Homeless 1"))
```

# Difference in means (e.g., the Age variable)

```{r}
dta_m %>% group_by(homeless) %>% select(one_of(c("age"))) %>%
  summarise_all(mean)

dta_m %>% group_by(homeless) %>% summarize(meanAge = mean(age)) 
```

```{r}
tapply(X = dta_m$age, INDEX = dta_m$homeless, FUN = mean)
```

# Estimating treatment effects

Estimating the risk effect is simple once we have a matched sample that we are
happy with.

```{r}
fit1 <- lm(cesd ~ homeless, data = dta_m, weights = weights)
```

# Cluster-robust standard errors

```{r}
library(lmtest)
coeftest(fit1,  cluster = ~subclass)
```
