---
title: "Mediation Analysis"
author: "BIOS 620"
date: "5/6/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
# install.packages("mediation")
library(mediation)
data("framing")
head(framing)
```

```{r,eval=FALSE}
?framing
```

## Model fits

$$
\begin{aligned}
  M_{}  = \alpha_{2} + \beta_{2}X_{} + \xi_{2}C_{} + \epsilon_{2} \quad &(2) \quad \text{mediator model} \\
\end{aligned}
$$

```{r}
med.fit <- lm(emo ~ treat + age + educ + gender + income, 
              data = framing)
```

$$
\begin{aligned}
  Y_{}  = \alpha_{3} + \beta_{3}X_{} + \gamma M_{} + \xi_{3}C_{}  + \epsilon_{3} \quad &(3) \quad \text{classic combined model} \\
\end{aligned}
$$

```{r}
out.fit <- glm(cong_mesg ~ treat + emo + age + educ 
               + gender + income, data = framing, 
               family = binomial)
```

## Mediation Analysis Results

$$
\begin{aligned}
   \text{ACME}(x) &= \text{E}(\delta_{i}(x)) \equiv \text{E}[Y_{i}(x, M_{i}(1)) - Y_{i}(x, M_{i}(0))],\\
   \text{ADE}(x) &= \text{E}(\phi_{i}(x)) \equiv \text{E}[Y_{i}(1, M_{i}(x)) - Y_{i}(0, M_{i}(x))] \\
\end{aligned}
$$  

```{r}
med.out <- mediate(med.fit, out.fit, 
                   treat = "treat", mediator = "emo", 
                   boot = TRUE, sims = 1000)
summary(med.out)
```   

# IRIS example with simulation

Follow this link to get the full post:
https://towardsdatascience.com/doing-and-reporting-your-first-mediation-analysis-in-r-2fe423b92171

```{r}
data("iris")
df <- iris
```

First, let's simulate the mediator, "attractiveness to the bee." This variable
will be named mediator and - for our example - will consist of two parts. 35% of
its value is Sepal.Length + 65% of its value is random noise.

```{r}
set.seed(12334)
df$random1 <- runif( nrow(df), min = min(df$Sepal.Length), 
                     max = max(df$Sepal.Length) )

df$mediator <- df$Sepal.Length * 0.35 + df$random1 * 0.65
```


simulate the dependent variable, "likelihood to be pollinated by a bee." This 
variable will be named dv and consists of two parts again. 35% of its value is
the Mediator, + 65% of its value is random noise. 

In our example, the random noise could be non-plant-specific attributes like the
location of the flower, weather conditions, soil, etc.

```{r}
df$random2 <- runif( nrow(df), min = min(df$mediator), max = max(df$mediator))
df$dv <- df$mediator * 0.35 + df$random2 * 0.65
```

## Step 1

Model with outcome and 'treatment'

```{r}
fit.totaleffect <- lm(dv ~ Sepal.Length, data = df)
summary(fit.totaleffect)
```

## Step 2

Mediator as the outcome variable in the model

```{r}
fit.mediator = lm(mediator ~ Sepal.Length, df)
summary(fit.mediator)
```

## Step 3

Full model (includes the mediator)

```{r}
fit.dv = lm( dv ~ Sepal.Length + mediator, df)
summary(fit.dv)
```

## Step 4

Use the `mediation` package:

```{r}
results <- mediate( fit.mediator, fit.dv, treat='Sepal.Length',
    mediator='mediator', boot=TRUE, sims = 1000)
summary(results)
```

# Lab Exercise

Re-design the simulation so that: 

+ a certain portion of mediator and a certain portion of dependent variable is
    coming from a confounder, e.g., the "petal.width" variable in the iris data.
+ re-run the mediation analysis 
    + (1) using the Matti Fichs's mediator model and outcome model (aka,
        mis-specified model)
    + (2) using the correct model where the confounder is added
+ compare (1) and (2) and comment results, upload to BB.


