---
title: "Multiple Imputation"
author: "BIOS 620"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup CHAIN dataset

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(haven)
```

## Read-in

```{r}
chain <- read_spss(file = "data/CHAIN.sav")
```


## Data inspection

```{r}
dim(chain)
head(chain)
```

### Variable names

* **log_virus** log of self reported viral load level
* **age** age at time of the interview
* **income** annual family income in 10 intervals
* **healthy** a continuous scale of physical health with a theoretical range
between 0 and 100 where better health is associated with higher scale values
* **mental** a binary measure of poor mental health ( 1=Yes, 0=No )
* **damage** ordered interval for the CD4 count, which is an indicator of how
much damage HIV (aka, the virus) has caused to the immune system
* **treatment** a three-level ordered variable: 0=Not currently taking HAART
(Highly Active AntiretRoviral Therapy) 1=taking HAART but nonadherent, 2=taking
HAART and adherent

# Model Fitting

## unadjusted model, i.e., univariate model

```{r}
univ.fit <- lm(log_virus ~ factor(treatment), data = chain)
```

## adjusted model, multivariate model

```{r}
mult.fit <- lm(log_virus ~ factor(treatment) + age + income + healthy + mental,
               data = chain)
```

### Comparing models

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(stargazer)
```

```{r}
stargazer(univ.fit, mult.fit,
    title="Results from univariate model (1) and multivariate model (2)",
    type="text",intercept.bottom=FALSE,
    style="default", omit.stat = c("rsq", "ser", "f", "adj.rsq"), header=FALSE)
```

## Missing data pattern 

```{r}
library(mice)
md.pattern(chain, plot=FALSE)
md.pattern(chain, plot=TRUE)
```

```{r}
chain$mental    <- factor(chain$mental)
chain$damage    <- factor(chain$damage)
chain$treatment <- factor(chain$treatment)
chain
```


# Impute the data

```{r}
names(chain)
head(chain, 3)
data.mice <- mice(chain, m = 50, 
    method = c("","pmm","pmm","pmm", "pmm", "logreg", "polr", "polr"),
    printFlag=TRUE)
```

## Fit model with imputed data

```{r}
data.mice
mice.fit <- with(data.mice,
    lm(log_virus ~ treatment + age  + income  + healthy  + mental)
)
```

## Get estimates with imputed data

```{r}
mi_result <-  pool(mice.fit)
```

## Compare imputed estimates vs multivar results 

```{r}
summary(mi_result)
summary(mult.fit)
```

