---
title: "Lab 2"
author: "BIOS 620"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load example data

## mtcars

Load the internal dataset provided with R:

```{r}
data(mtcars)
```

## Data inspection

```{r}
?mtcars
dim(mtcars)
head(mtcars)
```

# Data Export

You can use a number of functions depending on the 
file type that you want to use. For example,
to create a CSV file, use `write.csv` with a `data.frame`.

```{r}
write.csv(mtcars, file = "data/mtcars.csv", row.names = TRUE)
```

# Example Re-import 

We can then read in the data using read.csv. Remember this
is all happening within our working directory. See `getwd()`.

```{r}
dat <- read.csv("data/mtcars.csv", header = TRUE, row.names = 1)
head(dat)
```

# Visual Exploration

## Histogram

Distribution of a single continuous variable:

```{r}
hist(mtcars$mpg, main = "Histogram of MPG")
```

```{r}
boxplot(mtcars$mpg)
mean(mtcars$mpg)
```


## Boxplots

Grouped distributions of displacement by cylinder type:

```{r}
boxplot(disp ~ cyl, data = mtcars)
```

## Bi-variable comparison

### Scatterplot

```{r}
plot(mtcars$mpg, mtcars$wt)
```


## Multi-variable comparisons

```{r}
pairs(mtcars, main = "mtcars data", gap = .5)
```

```{r}
library(psych)
pairs.panels(mtcars)
```

# Simple Linear Regression

## Formula notation

A way to tell R how variables are related. Usually as
'Y = X1 + X2' except that the '=' is replaced with the
tilde ' ~ '. 

```{r}
mpg ~ wt
mpg ~ 1 + wt
```

Here `mpg` is the Y variable and `wt` is the X. 

## Model fit

Now we fit the linear model and see the fit object called
`myfit`.

```{r}
myfit <- lm(mpg ~ 1 + wt, data = mtcars)
myfit
myfit <- lm(mpg ~ wt, data = mtcars)
myfit
summary(myfit)
```

```{r}
plot(mtcars$wt, mtcars$mpg)
abline(
  # model 1
  lm(mpg ~ 1 + wt, data = mtcars),
  col = "blue", lwd = 2
)
## centering the weight variable
plot(mtcars$wt - mean(mtcars$wt), mtcars$mpg)
abline(
  # centering
  lm(mpg ~ 1 + I(wt - mean(wt)), data = mtcars),
  col = "red", lwd = 2
)
```


## Fitted values

```{r}
yhat <- predict(myfit) 
# actual vs fitted values
plot(mtcars$mpg, yhat)
abline(a=0,b=1)
```

# Reading a SAS file

```{r}
library(haven)
french <- read_sas("data/French_males.sas7bdat")
head(french)
```

