---
title: "Multivariate Regression for Confounding Models"
author: "BIOS 620 / PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulating Data

## Setting a seed

When running a simulation, it is important to set a random seed before
invoking a command that requires random number generation such as
`rnorm`. This keeps our results reproducible. A computer cannot generate
_truly_ random events.

```{r}
set.seed(1)
AGE <- rnorm(100, mean=35, sd=5)
SEX <- rbinom(100, 1, 0.5)
## SL is dependent on AGE and SEX
SL <-  rnorm(100, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
## True SL Coeff is 2.00
SBP <- rnorm(100, mean=109.47 + 2.00*SL + 0.5*AGE + 0.5*SEX, sd=5)
```

*Note*. Both AGE and SEX are confounders in this simulation and `SL` is
sugar level.

# Univariate analysis: between outcome and study variable

We fit a simple model. 

```{r}
fit1 <- lm(SBP ~ 1 + SL)
summary(fit1)
```

## Design matrix `model.matrix`

```{r}
head(model.matrix(SBP ~ 1 + SL))
head(model.matrix(SBP ~ SL))
```

# Multivariate analysis: adjusting confounder AGE 

*Note*. Here we are treating SEX as an "unobserved confounder" in this analysis by
intentionally omitting it.

```{r}
fit2 <- lm(SBP ~ 1 + SL + AGE)
summary(fit2)
```

## Extracting information from a fit object

```{r}
## coefficients
coef(fit2)
## conf. intervals
confint(fit2)
```

# Using the stargazer package for nice summary tables 

```{r}
library(stargazer)
# install.packages("stargazer")
stargazer(fit1, fit2, title="Regression Results",
    type="text", style="default", header=FALSE)
```

# Introduction to `for` loops

A simple example. Iterating over `i` values. The `length` of `i` indicates
the number of iterations. 

```{r}
i <- 1:3
length(i)
```

Note. `i` in the `.GlobalEnv` is different from the i *INSIDE* the `for` loop.

```{r}
for (i in 1:3) {
    print(i)
}
```

## Allocate and fill method

The recommended method for writing a `for` loop is to use the "allocate and
fill" method. It provides an efficient use of memory rather than the
"copy and append" method. First we create an empty vector or list.

```{r}
fill_vector <- vector(mode = "numeric", length = 10)
fill_vector
```

We then use the `for` loop to fill the values of that vector at each
iteration.

```{r}
for (i in 1:10) {
    fill_vector[i] <- i + 1
}
fill_vector
```

## Replicate

A generalized function for repeating the same expression with random
number generation. An 'expression' is any R language object. It is one of the
'parts of speech' in the R language. To encapsulate an 'expression' (the
operations that we want to repeat over a series of random numbers), we use the
open and close `{ }` curly braces.

```{r}
set.seed(1)
result <- replicate(6, { rnorm(n = 10, mean = 0, sd = 1) })
## inspect the class
class(result)
## see the object
result
```

# Lab exercise

1. set different seeds to see how the results changes, especially
    the coefficient for SL (sugar level)
    
```{r}


```

2. Choose different magnitude of confounding from the unobserved 
   confounder SEX, re-fit the model and observe the change of the
   impact of SL to SBP (the main study association)
   
```{r}


```

3. Try other changes you might consider, e.g., magnitude of error (for
   SL and/or SBP)
   
```{r}


```
   
Finally, write a `for` loop to gather results from 1000 seeds and
quantify the percent bias (PRB) of estimating coefficient of SL due to
omitting confounders.

$PRB = \frac{1}{S} \sum_{s=1}^{1000} 100 * \frac{\beta_1^{(s)}-\beta_1^{tru}}{\beta_1^{tru}}$

```{r}


```
