---
title: "Multivariate Regression for Confounding Models"
author: "BIOS 620 / PUBH 614"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab exercise

1. set different seeds to see how the results changes, especially
    the coefficient for SL (sugar level)
    
```{r}
###------ Change:  seed
myseed <- 10
set.seed(myseed)
AGE <- rnorm(100, mean=35, sd=5)
SEX <- rbinom(100, size = 1, prob = 0.5)
SL <-  rnorm(100, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
SBP <- rnorm(100, mean=109.47 + 2.00*SL + 0.5*AGE + 0.5*SEX, sd=5)

fit3 <- lm(SBP~1 + SL + AGE)
summary(fit3)

stargazer(fit1, fit2, fit3, title="Regression Results",
   type="text", style="default", header=FALSE)
```

2. Choose different magnitude of confounding from the unobserved 
   confounder SEX, re-fit the model and observe the change of the
   impact of SL to SBP (the main study association)
   
```{r}
beta2 <- 0.8
sdv <- 5
myseed <- 10
set.seed(myseed)

AGE <- rnorm(100, mean=35, sd=5)
SEX <- rbinom(100, size = 1, prob = 0.5)
SL <-  rnorm(100, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
SBP <- rnorm(100, mean=109.47 + 2.00 *SL + beta2*AGE + 0.5*SEX, sd = sdv)

fit1 <- lm(SBP~1 + SL + AGE)

#### -- change beta2
beta2 <- 0.2
SBP <- rnorm(100, mean=109.47 + 2.00 *SL + beta2*AGE + 0.5*SEX, sd= sdv)

fit2 <- lm(SBP~1 + SL + AGE)

stargazer(fit1, fit2, title="Regression Results",
   type="text", style="default", header=FALSE)
```

3. Try other changes you might consider, e.g., magnitude of error (for
   SL and/or SBP)
   
```{r}
###------ Change: beta1, beta2 and sdv
beta1 <- 1.5
beta2 <- 1
sdv <- 10
myseed <- 1
set.seed( myseed)
AGE <- rnorm(1000, mean=35, sd=5)
SEX <- rbinom(1000, size = 1, prob = 0.5)
SL <-  rnorm(1000, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
SBP <- rnorm(1000, mean=109.47 + beta1 *SL + beta2*AGE + 0.5*SEX, sd = sdv)

fit1 <- lm(SBP~1 + SL + AGE)
summary(fit1)
```
   
Finally, write a `for` loop to gather results from 1000 seeds / replications and
quantify the percent bias (PRB) of estimating coefficient of SL due to
omitting confounders.

$PRB = \frac{1}{S} \sum_{s=1}^{1000} 100 \times \frac{\beta_1^{(s)}-\beta_1^{tru}}{\beta_1^{tru}}$

```{r}
## answer with for loop
beta1.true <- 2.0
num_reps <- 1000
beta1.c <- vector("numeric", length = num_reps)

for (myseed in seq_len(num_reps)) {
   set.seed(myseed)
   AGE <- rnorm(1000, mean=35, sd=5)
   SEX <- rbinom(1000, size = 1, prob = 0.5)
   SL <-  rnorm(1000, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
   SBP <- rnorm(1000, mean=109.47 + beta1.true *SL + 0.5*AGE + 0.5*SEX, sd= 5)
   
   myfit <- lm(SBP~1 + SL + AGE + SEX)
   summary(myfit)
   beta1.c[myseed] <- coefficients(myfit)["SL"]
}

mean((beta1.c - beta1.true) / beta1.true) * 100

## answer with replicate
set.seed(2021)
beta1.c <- replicate(1000, {
   AGE <- rnorm(1000, mean=35, sd=5)
   SEX <- rbinom(1000, size = 1, prob = 0.5)
   SL <-  rnorm(1000, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
   SBP <- rnorm(1000, mean=109.47 + beta1.true *SL + 0.5*AGE + 0.5*SEX, sd= 5)
   
   myfit <- lm(SBP~1 + SL + AGE + SEX)
   summary(myfit)
   coef(myfit)[["SL"]]
})

mean((beta1.c - beta1.true) / beta1.true) * 100
```
