---
title: "General FA-Questions"
author: "BIOS 620"
date: "2/22/2021"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to import a data set?

1. Find your dataset after download.

### absolute path 

Use this function to find the _absolute path_ to your file. 

```{r,eval=FALSE}
file.choose()
# [1] "/cloud/project/data/birthwt.csv"
```

### relative path

Tip: It is a good idea to always put the data in the same folder as the
RMarkdown file that you are knitting.

Note that `getwd()` is good for R scripts but does not work for RMarkdown.

```{r}
# put the data where the Rmd lives
head( read.csv("birthwt.csv") )
```

2. Figure out what function do we need to use to import (file extension).

```{r}
# ?read.csv
birthwt <- read.csv(file = "birthwt.csv", header = TRUE, sep = ",")
# explore the data
head(birthwt)
dim(birthwt)
class(birthwt)
```

## What is y hat? 

```{r}
data("mtcars")
head(mtcars)

fit <- lm(mpg ~ wt, data = mtcars)

plot(mtcars$wt, mtcars$mpg)
abline(fit, col = "red")

# Yhat = A + B*X1
yhat_vals <- coef(fit)[1] + coef(fit)[2]*mtcars$wt
head(yhat_vals)
head(predict(fit))
```

## What do we need `rnorm` for?

```{r}
AGE <- rnorm(100, mean = 30, sd = 5)
```

