---
title: "Frequently Asked Questions"
author: "BIOS 620"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to import a data set?

1. Finding your dataset after download with either:

* file.choose
* Operating systems file explorer (or finder)

### file.choose

Use `file.choose()` function interactively to get the _absolute path_ to your
file.

```{r,eval=FALSE}
file.choose()
# [1] "/cloud/project/data/birthwt.csv"
```

Note. Windows users may have double backslashes in their paths!

### File explorer (finder)

* Use Shift + Right-click on Windows > Copy as Path
* File properties in the Finder for Mac

### Absolute paths

* Include the drive letter
* Are generally much longer than relative paths
* The return of `file.choose()` is an _absolute_ path.

### relative path

**Tip**: It is a good idea to always put the data in the same folder as the
RMarkdown file that you are knitting.

Note that `getwd()` is good for interactive use but does not work for
RMarkdown.

```{r}
# put the data where the Rmd lives
head( read.csv("birthwt.csv") )
```

### Actually Importing

Figure out what function do we need to use to import (based on the file
extension).

```{r}
# ?read.csv
birthwt <- read.csv(file = "birthwt.csv", header = TRUE, sep = ",")
# explore the data
head(birthwt)
dim(birthwt)
class(birthwt)
```

For SAS datasets, we recommend using the `haven` package.

```{r}
library(haven)
french <- read_sas("French_males.sas7bdat")
head(french)
```

## What is y hat?

```{r}
data("mtcars")
head(mtcars)

fit <- lm(mpg ~ wt, data = mtcars)

plot(mtcars$wt, mtcars$mpg)
abline(fit, col = "red")

# Yhat = A + B*X1
yhat_vals <- coef(fit)[1] + coef(fit)[2]*mtcars$wt
head(yhat_vals)
head(predict(fit))
```

## What do we need `rnorm` for?

`rnorm` and `rbinom` allow us to create a random variable from the specified
distribution (either normal or binomial in this case). Here we take 100 random
values from the normal distribution with a mean of 30 and a standard deviation
of 5:

```{r}
AGE <- rnorm(100, mean = 30, sd = 5)
```

