---
title: "Multivariate Regression for Confounding Models"
author: "BIOS 620 / PUBH 614"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulating Data

In order to understand the influence of unobserved confounders, we will create
a simulation study where we create random variables with _known_ population
parameters.

## Setting a seed

Because we are using a computer based on zeros and ones, _true_ randomness is
not possible. Therefore, we need a way to obtain "random" numbers with
an algorithm that attempts to do that.

We use the `set.seed` function to generate random numbers from an origin
'seed'. The 'seed' is an input number for that algorithm. We re-use the
'seed' to keep our simulation results shareable and reproducible. It affects
how functions that generate random numbers behave (such as `rnorm`).

```{r}
set.seed(1)
AGE <- rnorm(100, mean=35, sd=5)
SEX <- rbinom(100, 1, 0.5)
## SL is dependent on AGE and SEX
SL <-  rnorm(100, mean=50 + 0.01*AGE - 0.01*SEX, sd=2)
## True SL Coeff is 2.00
SBP <- rnorm(100, mean=109.47 + 2.00*SL + 0.5*AGE + 0.5*SEX, sd=5)
```

*Note*. Both AGE and SEX are confounders in this simulation and `SL` is
sugar level.

# Univariate analysis: between outcome and study variable

We fit a simple linear model.

```{r}
fit1 <- lm(SBP ~ 1 + SL)
summary(fit1)
```

## Side note: Design matrix with `model.matrix`

To verify the effect of the constant (1) in the R formula `SPB ~ 1 + SL`, we
use the `model.matrix` function to obtain the design matrix that R uses
to calculate regression estimates. By comparing the output of both
`model.matrix operations below, we see that the constant has no effect in how
the `model.matrix` is generated. It is included for illustrate the intercept
(`A`) in the equation

$ SBP = A + B_{1} * SL + B_{2} * AGE + E $

```{r}
head(model.matrix(SBP ~ 1 + SL))
head(model.matrix(SBP ~ SL))
```

# Multivariate analysis: adjusting confounder AGE

*Note*. Here we are treating SEX as an "unobserved confounder" in this analysis
by intentionally omitting it.

```{r}
fit2 <- lm(SBP ~ 1 + SL + AGE)
summary(fit2)
```

## Extracting information from a fit object

In instances where we need to compute on the output of a regression,
there are several functions to help us obtain pieces of the results.
For example, `coef` to obtain the coefficients and `confint` to obtain
the 95% confidence intervals of the estimates.

```{r}
## coefficients
coef(fit2)
## conf. intervals
confint(fit2)
```

# Using the stargazer package for nice summary tables

Use the `stargazer` package to obtain a nice output of your results.

```{r}
library(stargazer)
# install.packages("stargazer")
stargazer(fit1, fit2, title="Regression Results",
    type="text", style="default", header=FALSE)
```

# Introduction to `for` loops

We commonly use `i` as the iterating variable, within a `for` loop.
As practice, create a vector from 1 to 3 and then ask what the `length`
of that vector is with the `length` function.

```{r}
i <- 1:3
length(i)
```

Next, let's create a `for` loop where we tell R to print the `i` variable at
each cycle.

*Note*. Variable `i` in the `.GlobalEnv` (see it with `ls()`) and is different
from the `i` *INSIDE* the `for` loop.

```{r}
for (i in 1:3) {
    print(i)
}

for (i in letters) {
   print(paste("This is letter", i))
}

paste("This is letter", letters)
```

## Allocate and fill method

The recommended method for writing a `for` loop is to use the "allocate and
fill" method. It provides an efficient use of memory rather than the
"copy and append" method. First we create an empty vector or list.

```{r}
fill_vector <- vector(mode = "numeric", length = 10)
# fill_vector <- rep(x = 0, times = 10)
fill_vector
```

We then use the `for` loop to fill the values of that vector at each
iteration.

```{r}
for (i in 1:10) {
    fill_vector[i] <- i + 1
}
fill_vector
```

## Replicate

A generalized function for repeating the same expression with random
number generation. An 'expression' is any R language object. It is one of the
'parts of speech' in the R language. To encapsulate an 'expression' (the
operations that we want to repeat over a series of random numbers), we use the
open and close `{ }` curly braces.

```{r}
set.seed(1)
result <- replicate(6, { rnorm(n = 10, mean = 0, sd = 1) })
## inspect the class
class(result)
## see the object
result
```

# Lab exercise

1. set different seeds to see how the results changes, especially
    the coefficient for SL (sugar level)

```{r}


```

2. Choose different magnitude of confounding from the unobserved
   confounder SEX, re-fit the model and observe the change of the
   impact of SL to SBP (the main study association)

```{r}


```

3. Try other changes you might consider, e.g., magnitude of error (for
   SL and/or SBP)

```{r}


```

Finally, write a `for` loop to gather results from 1000 seeds and
quantify the percent bias (PRB) of estimating coefficient of SL due to
omitting confounders.

$PRB = \frac{1}{S} \sum_{s=1}^{1000} 100 \times \frac{\beta_1^{(s)}-\beta_1^{tru}}{\beta_1^{tru}}$

```{r}


```
