---
title: "Linear Regression with Categorical Variables"
author: "BIOS 620 / PUBH 614"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example Dataset: `mtcars`

## Motor Trend Car Road Tests

```{r}
data("mtcars")
```

Explore the data with some usual functions:

```{r}
# ?mtcars
dim(mtcars)
head(mtcars)
names(mtcars)
class(mtcars)
```

# Describe each variable numerically and visually

To describe a numeric variable, `summary` is a good function to use:

```{r}
summary(mtcars$mpg)
sd(mtcars$mpg)
```

To visualize the distribution of a numeric continuous variable, use
`hist` to see a histogram.

```{r}
hist(mtcars$mpg, main = "Histogram of mpg")
```

## Transmission

The transmission variable `am` is coded as `0` for automatic and `1` for
manual.

```{r}
table(mtcars$am)
prop.table( table(mtcars$am) )
```

To see how transmission type relates to miles per gallon, we use a simple
linear model:

*Note*. In R, the lowest value is the reference category by default.

```{r}
fit1 <- lm(mpg ~ 1 + am, data = mtcars)
summary(fit1)
```

To further verify, we can also look at the means in each group:

```{r}
tapply(X = mtcars$mpg, INDEX = mtcars$am, FUN = mean)
```

### Reverse dummy coding

Say we wanted to see the effect of having an automatic transmission compared
to a manual one, we would reverse the coding of the variable. We use
the `ifelse` function to create a conditional statement (i.e., does `am` equal
`0`?) to reverse the coding by setting `0` to `1` and all others (`1`) to `0`.

```{r}
(  mtcars$am_auto <- ifelse(mtcars$am == 0, 1, 0)  )
# mtcars$am_auto
```

We then fit the model again with our new variable `am.new`:

```{r}
fit2 <- lm(mpg ~ 1 + factor(am_auto), data = mtcars)
summary(fit2)
```

Note that we have to tell R that we have a factor variable by enclosing the
variable name in `factor()`, otherwise R will think it is a numeric variable.

We can now check the difference in the coefficients with these two coding
schemes:

```{r}
fit1
fit2
```

# Factors

The formal way to represent categorical variables in R is to use factors.
This avoids us the creation of dummy variables that is necessary in other
analysis software such as SPSS and SAS (for linear regressions).

To create a factor, we need to use a numeric vector variable (`am`) and
use the `factor` function. We can also add labels to each 'level' (unique
value) in the variable with the `labels` argument.

```{r}
am_factor <- factor(mtcars$am, levels = c(0, 1),
    labels = c("automatic", "manual"))
am_factor
```

We can then add our factor variable back into the data:

```{r}
mtcars$am_factor <- am_factor
head(mtcars)
```

## contrasts and relevel

In order to see the reference category for any factor, we can use
the contrasts function. It will show us a binary matrix (see lecture) where
the row with all zeros across is designated the reference category. Let's try
it on the `am_factor` variable:

```{r}
contrasts(am_factor)
```

When using factors in a linear model, R will take the lowest value
as the reference category. In order to change this, we can use the
`relevel` function.

```{r}
relevel(mtcars$am_factor, ref = "manual")
```

Note that we need to save this change back into the data for the changes
to hold:

```{r}
mtcars$am_factor <- relevel(mtcars$am_factor, ref = "manual")
```

Check that the reference category has changed:

```{r}
contrasts(mtcars$am_factor)
```

### Practice

Run `contrasts` on the cylinder (`cyl`) variable in the `mtcars` dataset:

```{r}

```

# Describe pairwise relationships

To visually compare `mpg` to `am` (transmission), we could use a `boxplot`.

```{r}
boxplot(mpg ~ am, data = mtcars)
```

We can also visually compare all the variables in the data:

```{r}
library(psych)
pairs.panels(mtcars)
```

We can also describe the relationship between `mpg` and `am` numerically:

```{r}
with(mtcars, cor(am, mpg))
# equivalently
# cor(mtcars$am, mtcars$mpg)
```

# Reading an SPSS file

SPSS files end in the `.sav` file extension. To import these type of files
into R, we recommend using the `haven` package. Remember, load the package
first with the `library` function.

```{r}
library(haven)
nhanes <- read_spss("NHANES_Lab4.sav")
```

Note that the NHANES dataset comes with labels:

```{r}
(edu_levels <- attributes(nhanes$Education)$labels)
```

We can use these labels to create a factor:

```{r}
factor(nhanes$Education, levels = edu_levels, labels = names(edu_levels))
```

# Exercises

Using the National Health and Nutrition Examination Survey (NHANES) data (SPSS
`.sav` format) to fit a multivariate model with `BPXSY1` as response variable,
`Education` as study variable, `Exactage` and `Male` as confounders.

* describe each variable numerically and visually

```{r}

```

* describe pairwise relationship between the study variable and outcome
variable numerically and visually (no hypothesis testing needed)

```{r}

```

* for the "Education" variable (which is a categorical variable), fit the model
in two ways and compare the results:
    + creating two dummy variables using if/else condition
    + using the factor() function
      + Define/change reference category for categorical variable in R

```{r}

```
