---
title: "Linear Regression with Categorical Variables"
author: "BIOS 620 / PUBH 614"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

Using the National Health and Nutrition Examination Survey (NHANES) data (SPSS
`.sav` format) to fit a multivariate model with `BPXSY1` as response variable,
`Education` as study variable, `Exactage` and `Male` as confounders.

* describe each variable numerically and visually

First, read in the data:

```{r}
library(haven)
nhanes <- read_spss("NHANES_Lab4.sav")
```

## Summarize and visualize numeric variables

### Systolic Blood Pressure

```{r}
summary(nhanes$BPXSY1)
hist(nhanes$BPXSY1)
boxplot(nhanes$BPXSY1)
```

### Exact age

```{r}
summary(nhanes$Exactage)
hist(nhanes$Exactage)
boxplot(nhanes$Exactage)
```

## Summarize and visualize categorical variables

### Sex

We can leave `Male` as-is because it is binary.

```{r}
table(nhanes$Male)
prop.table( table(nhanes$Male) )
```

### Education

*Note*. We want to create an `Education` factor with labels for informative 
plots and outputs.

```{r}
table(nhanes$Male)
edu_levels <- attributes(nhanes$Education)$labels
nhanes$Edu_f <- factor(nhanes$Education, levels = edu_levels,
    labels = names(edu_levels))
table(nhanes$Edu_f)
prop.table( table(nhanes$Edu_f) )
```

## Describe pairwise relationships

* between the study and outcome variables  numerically and visually (no
hypothesis testing needed)

Mean SBP by Education level

```{r}
tapply(X = nhanes$BPXSY1, INDEX = nhanes$Edu_f, FUN = mean, na.rm = TRUE)
## or
aggregate(x = nhanes$BPXSY1, by = list(Edu_level = nhanes$Edu_f),
    FUN = mean, na.rm = TRUE)
## or even
sapply(split(nhanes$BPXSY1, nhanes$Edu_f), mean, na.rm = TRUE)
```

We add proper labels, our outputs are easily understood. We can also add
comments and descriptions in this our RMarkdown document. 

### Correlation

Spearman correlation (is there a non-parametric monotonic relationship?)

```{r}
with(nhanes,
     cor(Education, BPXSY1, use = "pairwise.complete.obs", method = "spearman")
)
```

Visualize SBP distribution by Education level:

```{r}
boxplot(BPXSY1 ~ Edu_f, data = nhanes)
```

## Fit the model

* for the "Education" variable (which is a categorical variable), fit the model
in two ways and compare the results:
    + creating two dummy variables using if/else condition
    + using the factor() function
      + Define/change reference category for categorical variable in R

### Creating two dummy variables

First, we reference our categories to guide the dummy recoding. 

```{r}
edu_levels
```

*Note*. Because we want the direction of the effect (coefficient) to be
positive, in this case, we use "Graduated from College" as the reference.

```{r}
nhanes$No_HS     <- ifelse(test = nhanes$Education == 1,  yes = 1, no = 0)
nhanes$HS <- ifelse(test = nhanes$Education == 2, yes = 1, no = 0)
fit1 <- lm(BPXSY1 ~ 1 + No_HS + HS + factor(Male) + Exactage, data = nhanes)
summary(fit1)
```

### Using the `factor` function

We can re-demo the factor creation from above because we are using a different
name for the factor variable (`Edu_f`).

```{r}
edu_levels <- attributes(nhanes$Education)$labels
nhanes$Edu_f <- factor(nhanes$Education, levels = edu_levels,
    labels = names(edu_levels))
```

We check the reference category in the variable:

```{r}
contrasts(nhanes$Edu_f)
```

And relevel so that "Graduated from College" is the reference.

```{r}
nhanes$Edu_f <- relevel(nhanes$Edu_f, ref = "Graduated from College")
```

We always check our work:

```{r}
contrasts(nhanes$Edu_f)
```

Now, we re-fit the model using this factor variable.

```{r}
fit2 <- lm(BPXSY1 ~ 1 + Edu_f + factor(Male) + Exactage, data = nhanes)
summary(fit2)
```

Positive coefficients!
