---
title: "OLS Regression Example"
author: "BIOS 620 / PUBH 614"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Visual Example of OLS Regression

Inspired by https://setosa.io/ev/ordinary-least-squares-regression/

## Example dataset `mtcars`

Take a subset of the `mtcars` dataset (cars less than 3k lbs.):

```{r}
data("mtcars", package = "datasets")
mtcars <- mtcars[mtcars$wt < 3, ]
```

# Simple Linear Regression Model

```{r,eval=FALSE}
lm(mpg ~ qsec, data = mtcars)
```


```{r tabsets, echo=FALSE}
suppressPackageStartupMessages({
  library(plotly)
  library(purrr)
  library(broom)
  library(shiny)
})

shinyApp(
  ui = fluidPage(
    verbatimTextOutput("summary"),
    plotlyOutput("p")
  ),

  server = function(input, output, session) {
    rv <- reactiveValues(
      mpg = mtcars$mpg,
      qsec = mtcars$qsec
    )
    grid <- reactive({
      data.frame(qsec = seq(min(rv$qsec) - 4, max(rv$qsec) + 4, length = 10))
    })
    model <- reactive({
      mtcars <- data.frame(qsec = rv$qsec, mpg = rv$mpg)
      lm(mpg ~ qsec, data = mtcars)
    })

    output$p <- renderPlotly({
      # creates a list of circle shapes from x/y data
      circles <- map2(rv$qsec, rv$mpg,
                      ~list(
                        type = "circle",
                        # anchor circles at (mpg, qsec)
                        xanchor = .x,
                        yanchor = .y,
                        # give each circle a 2 pixel diameter
                        x0 = -4, x1 = 4,
                        y0 = -4, y1 = 4,
                        xsizemode = "pixel",
                        ysizemode = "pixel",
                        # other visual properties
                        fillcolor = "blue",
                        line = list(color = "transparent")
                      )
      )
      squares <- Map(function(x0, x1, y0, y1) {
        list(
          type = "rect",
          fillcolor = "orange",
          line = list(color = "orange"),
          xref = "x",
          x0 = x0, x1 = x1,
          yref = "y",
          y0 = y0, y1 = y1,
          opacity = 0.2
        )
      }, x0 = rv$qsec, x1 = (rv$qsec - residuals(model())),
        y0 = predict(model()), y1 = rv$mpg)

      f <- list(
        family = "Courier New, monospace",
        size = 18,
        color = "#7f7f7f"
      )
      xlab <- list(
        title = "Quarter Mile Time (sec)",
        titlefont = f
      )
      ylab <- list(
        title = "Miles per Gallon (mpg)",
        titlefont = f,
        scaleanchor = "x",
        scaleratio = 1
      )

      # plot the shapes and fitted line
      plot_ly(width = 800, height = 800) %>%
        add_lines(x = grid()$qsec,
                  y = predict(model(), grid()), color = I("red")) %>%
        layout(
          shapes = c(circles, squares), xaxis = xlab, yaxis = ylab,
          autosize = FALSE
        ) %>%
        config(edits = list(shapePosition = TRUE)) %>%
        add_annotations(x = rv$qsec,
                  y = rv$mpg,
                  text = rownames(mtcars),
                  textangle = 45,
                  xref = "x",
                  yref = "y",
                  showarrow = TRUE,
                  arrowwidth = 1,
                  ax = 10,
                  ay = -20,
                  font = list(
                      color = '#264E86',
                      family = 'sans serif',
                      size = 14
                  )
        )
    })

    output$summary <- renderPrint({
      summary(model())
    })

    # update x/y reactive values in response to changes in shape anchors
    observe({
      ed <- event_data("plotly_relayout")
      shape_anchors <- ed[grepl("^shapes.*anchor$", names(ed))]
      if (length(shape_anchors) != 2) return()
      row_index <- unique(readr::parse_number(names(shape_anchors)) + 1)
      pts <- as.numeric(shape_anchors)
      rv$qsec[row_index] <- pts[1]
      rv$mpg[row_index] <- pts[2]
    })

  },
  options = list(height = 1200)
)
```

This example was adapted from `plotly::plotly_example("shiny", "drag_markers")`.

```{r,warning=FALSE,message=FALSE}
library(statsr)
```

Now, try running this interactively. Highlight the line of code and
run it in the console.

```{r,eval=FALSE, echo=TRUE}
plot_ss(qsec, mpg, mtcars, showSquares = TRUE)
```

Here we see the line using OLS:

```{r}
plot_ss(qsec, mpg, mtcars, showSquares = TRUE, leastSquares = TRUE)
```

We can also do this with the stock graphics engine in R:

```{r}
plot(mtcars$qsec, mtcars$mpg, asp = 1)
fit1 <- lm(mpg ~ qsec, data = mtcars)
abline(fit1, col = "blue")
# identify(mtcars$qsec, mtcars$mpg)
rownum <- 20
polygon(
  x = c(
    rep(mtcars$qsec[rownum], 2),
    rep(mtcars$qsec[rownum] - c(mtcars$mpg[rownum] - fitted(fit1)[rownum]), 2)
  ),
  y = c(
    fitted(fit1)[rownum], rep(mtcars$mpg[rownum], 2), fitted(fit1)[rownum]
  ),
  border = "red"
)
```
