---
title: "RCT / Review Session"
author: "BIOS 620"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Faraway Dataset

```{r}
library(faraway)
data(penicillin)
head(penicillin)
```

# Sidenote: Reshaping data (wide format)

```{r}
library(tidyr)
pivot_wider(penicillin, names_from = "treat", values_from = "yield")
```


## Penicillin yield based on blend and processes

```{r}
g <- lm(yield ~ treat+blend, data=penicillin)
```

```{r}
summary(g)
```

Here we fit the models that we will use to compare to the full (leave the
main effect variable out)

```{r}
fit0 <- lm(yield ~ treat, data = penicillin)
fit1 <- lm(yield ~ blend, data = penicillin)
```

Is the main effect of blend significant?

```{r}
anova(g, fit0)
```

Is the main effect of treatment significant?

```{r}
anova(g, fit1)
```

## Exploratory Data Analysis and Diagnostic Plots

```{r}
par(mfrow=c(2,2))
boxplot(yield~treat, data= penicillin, las=1)
boxplot(yield~blend, data= penicillin, las=1)
plot(predict(g), resid(g), xlab="Fitted", ylab="Residuals", las=1)
abline(h=0, lty=2)
qqnorm(resid(g), las=1)
qqline(resid(g), col=2)
```

# Visualizing contrasts

```{r}
suppressPackageStartupMessages(library(PairViz))
g2 <-  aov(yield ~ treat + blend, data= penicillin)
tuk <- TukeyHSD(g2,conf.level = 0.95)
par(mfrow=c(1,2))
par(mar=c(5,6,4,2))
plot(tuk, las=1, tcl = -.3)
```

# Review

```{r}
data(mtcars)
```

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
fit <- lm(mpg ~ 1 + wt, data = mtcars)
```

## Using the regression equation with new data

For example, we want to predict the `mpg` for a car that weighs 4000 lbs.

```{r}
predict(fit, newdata = data.frame(wt = 4))
```

```{r}
coefs <- unname(coef(fit))
intercept <- coefs[1]
betawt <- coefs[2]
intercept + (betawt*4)
```

## With multiple linear regresion

```{r}
library(haven)
nhanes <- read_spss("../data/NHANES_Lab4.sav")
```

### create our factor

```{r}
edu_levels <- attributes(nhanes$Education)$labels
nhanes$Education_f <- factor(nhanes$Education,
    levels = edu_levels, labels = names(edu_levels))
table(nhanes$Education_f)

# as_factor(nhanes$Education)
```

### Inspect our reference category...

`"Not graduated from HS"` is the reference:

```{r}
contrasts(nhanes$Education_f)
```

```{r}
fit <- lm(BPXSY1 ~ Education_f + Male + Exactage, data = nhanes)
summary(fit)
```

## Create a small data.frame with new data

```{r}
predict(fit,
    data.frame(Exactage = 30, Education_f = "Graduated from College", Male = 0)
)
```

### Manual calculation using the regression equation

```{r}
(model_coefs <- coefficients(fit)[-c(2, 4)])
# age * 30 + intercept + EduCollege * 1
(model_coefs[3]*30) + sum(model_coefs[1:2])
```

Alternatively, we can multiply by a vector of values that correspond to the
Xs in our new data. The zeros here mean that the Beta for that category does not
correspond to our new data (e.g., `"Graduated from HS"` $\beta$ * 0).

```{r}
coefficients(fit)
sum(coefficients(fit) * c(1, 0, 1, 0, 30))
```

Note. We just used a vectorized operation!

### Reference group changes

The aim here is to compare beta values when reference groups change.
The differences in magnitudes across the groups should be the same!

First, we two models where the only difference is the reference group
for education:

```{r}
contrasts(nhanes$Education_f)
nhanes$Education_f <- relevel(nhanes$Education_f, ref = "Not graduated from HS")
fit0 <- lm(BPXSY1 ~ Education_f + Male + Exactage, data = nhanes)

contrasts(nhanes$Education_f)
nhanes$Education_f <- relevel(nhanes$Education_f, ref = "Graduated from College")
fit1 <- lm(BPXSY1 ~ Education_f + Male + Exactage, data = nhanes)
```

We can have a look at the summary of both fit objects:

```{r}
# 'Not graduated from HS' is the reference
summary(fit0)
# 'Graduated from College' is the reference
summary(fit1)
```

To make a more direct comparison, we can construct some nice output for
both sets of coefficients for the Education variable and also include the
reference category:

```{r}
f0coeff <- c(`Not graduated from HS` = 0, coef(fit0)[2:3])
names(f0coeff) <- gsub("Education_f", "", names(f0coeff))
f0 <- f0coeff[order(names(f0coeff))]
```

```{r}
f1coeff <- c(`Graduated from College` = 0, coef(fit1)[2:3])
names(f1coeff) <- gsub("Education_f", "", names(f1coeff))
f1 <- f1coeff[order(names(f1coeff))]
```

Let's see both now:

```{r}
f0
f1
```

The coefficients are the same whether you add or subtract 5 depending on
the reference category that changes.


