---
title: "Logistic Regression"
author: "BIOS 620"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Help RCT

```{r}
library(haven)
help_rct <- read_spss("data/data_HELP.sav")
```

```{r}
with(help_rct, table(treat, linkstatus))
```

## Fit the logistic regression model


```{r}
fit <- glm(linkstatus ~ 1 + treat, family=binomial, data=help_rct)
fit <- glm(linkstatus ~ 1 + treat, family="binomial", data=help_rct)
```

## Beta Estimates

```{r}
summary(fit)
```

## Odds Ratio

```{r}
exp(fit$coefficient[2])
```

# Titanic Example

What is the probability of surviving the Titanic based on sex?

```{r}
library(titanic)
data(titanic_train)
titanic <- titanic_train
```

```{r}
table(titanic$Sex)
table(titanic$Sex, titanic$Survived)
```

```{r}
fit <- glm(Survived ~ 1 + Sex, family=binomial, data=titanic)
summary(fit)
```

```{r}
female <- factor(titanic$Sex, levels = c("male", "female"))
contrasts(female)

titanic$female <- relevel(female, ref = "male")

fit1 <- glm(Survived ~ female, family = binomial, data = titanic)
summary(fit1)
exp(coef(fit1))
```

```{r}
# ifelse(titanic$Sex == "female", 1, 0)
titanic$fem01 <- as.numeric(titanic$Sex == "female")
fit1 <- glm(Survived ~ fem01, family = binomial, data = titanic)
summary(fit1)
exp(coef(fit1))
```



## Odds Ratio

```{r}
exp(coef(fit))
```

## Confidence Intervals

```{r}
exp(confint(fit1))
```

## Manual calculation

\[
 \hat{p}_{i} = \frac{\exp (\hat{\beta}_{0} + \hat{\beta}_{1} x_{i})}{1 + \exp (\hat{\beta}_{0} + \hat{\beta}_{1} x_{i})}
\]

What is the probability of surviving as female?

```{r}
exp(coef(fit)[1]) / ( 1 + exp(coef(fit)[1]) )
```

## Predict function

```{r}
predict(fit, newdata = data.frame(Sex = "female"), type = "response")
```

## Side note (Mosaic plots)

```{r}
data("Titanic")
library(vcd)
mosaic(Titanic)
```


# Exercise

Calculate the probability of o-ring failure with the Challenger data at a
temperature of **31** degrees Fahrenheit. 

```{r}
library(readxl)
challenger <- read_excel("data/prechallenger.xls")
challenger
```

