---
title: "Logistic Regression"
author: "BIOS 620"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Help RCT

```{r}
library(haven)
help_rct <- read_spss("data/data_HELP.sav")
```

```{r}
with(help_rct, table(treat, linkstatus))
```

## Fit the logistic regression model


```{r}
fit <- glm(linkstatus ~ 1 + treat, family=binomial, data=help_rct)
fit <- glm(linkstatus ~ 1 + treat, family="binomial", data=help_rct)
```

## Beta Estimates

```{r}
summary(fit)
```

## Odds Ratio

```{r}
exp(fit$coefficients[2])
coef(fit)
coefficients(fit)
```

# Titanic Example

What is the probability of surviving the Titanic based on sex?

```{r}
library(titanic)
data(titanic_train)
titanic <- titanic_train
```

```{r}
table(titanic$Sex)
prop.table(table(titanic$Sex))
table(titanic$Sex, titanic$Survived)
```

```{r}
fit <- glm(Survived ~ 1 + Sex, family=binomial, data=titanic)
summary(fit)
```

Perhaps we want to model the probability of surviving as female.
We would have to change the reference group from `male` to `female`.

```{r}
sex_f <- factor(titanic$Sex, levels = c("male", "female"))
contrasts(sex_f)

with(titanic, table(Sex, sex_f))

titanic$female <- relevel(sex_f, ref = "male")

fit1 <- glm(Survived ~ sex_f, family = binomial, data = titanic)
summary(fit1)
exp(coef(fit1))
```

## Creating a dummy variable from a character one

```{r}
# ifelse(titanic$Sex == "female", 1, 0)
titanic$fem01 <- as.numeric(titanic$Sex == "female")
fit0 <- glm(Survived ~ fem01, family = binomial, data = titanic)
summary(fit0)
exp(coef(fit0))
```


## Odds Ratio

```{r}
exp(coef(fit1))
```

## Confidence Intervals

```{r}
exp(confint(fit1))
```

## Manual calculation

\[
 \hat{p}_{i} = \frac{\exp (\hat{\beta}_{0} + \hat{\beta}_{1} x_{i})}{1 + \exp (\hat{\beta}_{0} + \hat{\beta}_{1} x_{i})}
\]

What is the probability of surviving as female?

Because we have modeled `male` to be the reference, our calculation
would produce a probability of surviving for the `female` group.

```{r}
exp(sum(coef(fit1))) / ( 1 + exp(sum(coef(fit1))) )
```

## Predict function

Less manually, we can get the probability of surviving for `female` using the
`predict` function:

```{r}
predict(fit1, newdata = data.frame(sex_f = "female"), type = "response")
```

We can do the same for `male`:

```{r}
predict(fit1, newdata = data.frame(sex_f = "male"), type = "response")
```

How do we report Odds Ratios for continuous outcomes?

```{r}
fit2 <- glm(Survived ~ Age, family = binomial, data = titanic)
summary(fit2)
exp(coef(fit2)[2])
```

For every unit increase in age, the odds of surviving is multiplied by 0.99 
(indicating decreased odds of surviving).

You can change the scale of the Age variable to make it easier to interpret.
Divide the age variable by 10 to get an odds ratio for every _decade_ increase
in age.

```{r}
titanic$ageby10 <- titanic$Age/10
fit3 <- glm(Survived ~ ageby10, family = binomial, data = titanic)
summary(fit3)
exp(coef(fit3)[2])
```

For every decade increase in age, the odds of surviving is 0.89 times the
baseline odds.

You can look at the probability of survival for several decade values.

```{r}
predict(fit3, newdata = data.frame(ageby10 = 2), type = "response")
predict(fit3, newdata = data.frame(ageby10 = 3), type = "response")
predict(fit3, newdata = data.frame(ageby10 = 4), type = "response")
```

## Side note (Mosaic plots)

```{r}
data("Titanic")
library(vcd)
mosaic(Titanic)
```

# Exercise

Calculate the probability of o-ring failure with the Challenger data at a
temperature of **31** degrees Fahrenheit. 

```{r}
library(readxl)
challenger <- read_excel("data/prechallenger.xls")
challenger
```

