---
title: "Interaction"
author: "BIOS 620"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup NHANES dataset

```{r}
library(haven)
library(stargazer)
nhanes <- read_spss("data/NHANES_RN1.sav")
```

# Linear Regression
## Example 1 - SBP on Age & Gender (NHANES 2005-2006, age 20 - 84)

```{r}
fit1 <- lm(BPXSY1 ~ 1 + Exactage + Male, data=nhanes)
fit2 <- lm(BPXSY1 ~ 1 + Exactage + Male + Exactage:Male, data=nhanes)
# equivalenly
fit2 <- lm(BPXSY1 ~ 1 + Exactage + Male + Exactage * Male, data=nhanes)
```

## Visual inspection of the relationship between X and Y

```{r}
with(nhanes, plot(BPXSY1, Exactage))
```

```{r}
table(nhanes$Male)
```

```{r}
summary(fit1)
summary(fit2)
stargazer(fit1, fit2,
    title="Without and with interaction for linear regression",
    df=FALSE, intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

## Using complete.cases

```{r}
complete2 <- complete.cases(nhanes[, c("BPXSY1", "Exactage", "Male")])
nhanes2 <- nhanes[complete2, ]
```

## Visualizing results

```{r}
plot( nhanes2$Exactage, nhanes2$BPXSY1, xlab="Age", ylab="SBP", color='gray85', pch=".", las=1)
fit.male   <- lm(BPXSY1~1 + Exactage , data = subset(nhanes2, Male==1))
fit.female <- lm(BPXSY1~1 + Exactage,  data = subset(nhanes2, Male==0))
abline(fit.female, col = "red")
abline(fit.male, col = "blue")
legend("topleft", bty='n', c("Male", "Female"), lty=c(1,1), col=c("blue", "red"))
```

# Example 2: Hypertension on Gender and Education

```{r}
nhanes$Hypertension <- as.numeric(nhanes$BPXSY1 > 140)
```

## Using complete.cases to subset

```{r}
complete3 <- complete.cases(nhanes[, c("Hypertension", "Education", "Male")])
nhanes3 <- nhanes[complete3, ]
```

### create our factor

```{r}
nhanes3$Edu <- as_factor(nhanes3$Education)
```

## Testing for the interaction

```{r}
fit3 <- glm(Hypertension ~ 1 + Male + Edu, family='binomial', data=nhanes3)

fit4 <- glm(Hypertension ~ 1 + Male + Edu + Male:Edu, family='binomial', data=nhanes3)
```

## Results

```{r}
stargazer(fit3, fit4,
    title="Without and with interaction for logistic regression",
    df=FALSE, intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

## Analysis of Deviance (Likelihood Ratio Test)

```{r}
anova(fit3, fit4, test='Chisq')
```

```{r}
library(lmtest)
lrtest(fit3, fit4)
```

## Stratified analysis

```{r}
dat.noHS <- subset(nhanes3, Education == 1)
dat.HS   <- subset(nhanes3, Education == 2)
dat.College <- subset(nhanes3, Education == 3)
```

### Model fit

```{r}
fit5a <- glm(Hypertension~1 + Male,  family='binomial', data=dat.noHS)
fit5b <- glm(Hypertension~1 + Male,  family='binomial', data=dat.HS)
fit5c <- glm(Hypertension~1 + Male,  family='binomial', data=dat.College)
```

```{r}
stargazer(fit5a, fit5b, fit5c,
    title="Stratified analysis between gender and hypertension", df=FALSE,intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

### Addressing repetitive code

You can use a list representation and iterate over the list.
See also the `purrr` package in the `tidyverse`.

```{r}
nhaneslist <- split(nhanes3, nhanes3$Education)
res <- lapply(nhaneslist, function(model) {
    glm(Hypertension ~ 1 + Male, family = "binomial", data = model)
})
```

### Same results

```{r}
stargazer(res,
    title="Stratified analysis between gender and hypertension", df=FALSE,intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

# Common reference group

```{r}
contrasts(nhanes3$Edu)
contrasts(factor(nhanes3$Male))
```

## Female and Not Graduate from HS are the common reference group

```{r}
nhanes3$SexEdu <- with(nhanes3, interaction(Male, Edu))
```

```{r}
fit0 <- glm(Hypertension ~ SexEdu, family = "binomial", data = nhanes3)
summary(fit0)
```

## Additive vs Multiplicative interaction?

As a good epidemiologist, you should be reporting both.
This is an important EPI topic.

