---
title: "Interaction"
author: "BIOS 620"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup NHANES dataset

```{r}
library(haven)
nhanes <- read_spss("data/NHANES_RN1.sav")
```

# Q&A Warmup

Means by group:

```{r}
sort(names(nhanes))
with(nhanes, tapply(X = BPXSY1, INDEX = Male, FUN = mean, na.rm = TRUE))
tapply(nhanes$BPXSY1, nhanes$Male, FUN = mean, na.rm = TRUE)

mean(unlist(nhanes[nhanes$Male == 1, "BPXSY1"]), na.rm = TRUE)
mean(unlist(nhanes[nhanes$Male == 0, "BPXSY1"]), na.rm = TRUE)

library(dplyr)
group_by(nhanes, Male) %>% summarize(means = mean(BPXSY1, na.rm = TRUE))
```

```{r}
boxplot(BPXSY1 ~ Male, data = nhanes)
```



# Linear Regression
## Example 1 - SBP on Age & Gender (NHANES 2005-2006, age 20 - 84)

```{r}
fit1 <- lm(BPXSY1 ~ 1 + Exactage + Male, data=nhanes)
fit2 <- lm(BPXSY1 ~ 1 + Exactage + Male + Exactage:Male, data=nhanes)
# equivalenly
fit2 <- lm(BPXSY1 ~ 1 + Exactage + Male + Exactage * Male, data=nhanes)
fit2 <- lm(BPXSY1 ~ 1 + Exactage * Male, data=nhanes)
```

```{r}
head(model.matrix(BPXSY1 ~ 1 + Exactage + Male + Exactage:Male, data=nhanes))
head(model.matrix(BPXSY1 ~ 1 + Exactage + Male + Exactage * Male, data=nhanes))
```

## Visual inspection of the relationship between X and Y

```{r}
with(nhanes, plot(Exactage, BPXSY1))
```

```{r}
table(nhanes$Male)
```

```{r}
summary(fit1)
summary(fit2)
library(stargazer)
stargazer(fit1, fit2,
    title="Without and with interaction for linear regression",
    df=FALSE, intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

## Using complete.cases

```{r}
complete2 <- complete.cases(nhanes[, c("BPXSY1", "Exactage", "Male")])
table(complete2)
nhanes2 <- nhanes[complete2, ]
```

## Visualizing the interaction

```{r}
with(nhanes2,
    plot(Exactage, BPXSY1, xlab="Age", ylab="SBP", las=1)
)
# nhanes2[nhanes2$Male == 1, ]
fit.male   <- lm(BPXSY1 ~ 1 + Exactage , data = subset(nhanes2, Male==1))
fit.female <- lm(BPXSY1 ~ 1 + Exactage,  data = subset(nhanes2, Male==0))
abline(fit.female, col = "blue", lwd = 3)
abline(fit.male, col = "red", lwd = 3)
legend("topleft", bty="n", c("Male", "Female"), lty=c(1,1), col=c("red", "blue"))
```

# Example 2: Hypertension on Gender and Education

```{r}
nhanes$Hypertension <- as.numeric(nhanes$BPXSY1 > 140)
table(nhanes$Hypertension)
```

## Using complete.cases to subset

```{r}
complete3 <- complete.cases(nhanes[, c("Hypertension", "Education", "Male")])
table(complete3)
nhanes3 <- nhanes[complete3, ]
dim(nhanes3)
```

### create our factor

```{r}
nhanes3$Edu_f <- as_factor(nhanes3$Education)
```

## Testing for the interaction

```{r}
fit3 <- glm(Hypertension ~ 1 + Male + Edu_f, family='binomial', data=nhanes3)

fit4 <- glm(Hypertension ~ 1 + Male + Edu_f + Male:Edu_f, family='binomial', data=nhanes3)
```

## Results

```{r}
stargazer(fit3, fit4,
    title="Without and with interaction for logistic regression",
    df=FALSE, intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

```{r}
head(model.matrix(Hypertension ~ 1 + Male + Edu_f + Male:Edu_f, data=nhanes3))
```


## Analysis of Deviance (Likelihood Ratio Test)

```{r}
anova(fit3, fit4, test='Chisq')
```

```{r}
library(lmtest)
lrtest(fit3, fit4)
```

## Stratified analysis

```{r}
dat.noHS <- subset(nhanes3, Edu_f == "Not graduated from HS")
dat.HS   <- subset(nhanes3, Edu_f == "Graduated from HS (but not college)")
dat.College <- subset(nhanes3, Edu_f == "Graduated from College")
```

```{r}
sum(is.na(nhanes3$Edu_f))
```

### Model fit

```{r}
fit5a <- glm(Hypertension~1 + Male,  family='binomial', data=dat.noHS)
fit5b <- glm(Hypertension~1 + Male,  family='binomial', data=dat.HS)
fit5c <- glm(Hypertension~1 + Male,  family='binomial', data=dat.College)
```

```{r}
stargazer(fit5a, fit5b, fit5c,
    title="Stratified analysis between gender and hypertension", df=FALSE,intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

```{r}
library(purr)
mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .x)) %>%
  map(summary) %>%
  map_dbl("r.squared")
```

```{r}
nhanes3 %>%
  split(.$Edu_f) %>%
  map(~ glm(Hypertension ~ 1 + Male,  family='binomial', data= .x)) %>%
  map(summary)
```


### Addressing repetitive code

You can use a list representation and iterate over the list.
See also the `purrr` package in the `tidyverse`.

```{r}
nhaneslist <- split(nhanes3, nhanes3$Edu_f)
res <- lapply(nhaneslist, function(model) {
    glm(Hypertension ~ 1 + Male, family = "binomial", data = model)
})

summ <- lapply(nhaneslist, function(model) {
    summary(glm(Hypertension ~ 1 + Male, family = "binomial", data = model))
})
```

### Same results

```{r}
stargazer(res,
    title="Stratified analysis between gender and hypertension", df=FALSE,
    intercept.bottom=FALSE,
    type="text", omit.stat=c("all"),
    style="default", header=FALSE
)
```

# Estimates Table (NO INTERACTION)

|             | NoHS | HS | College |
|-------------|------|:--:|---------|
| Female      |      |    |         |
| Male        |      |    |         |
| Male-Female |      |    |         |

### Column 1

Male & Not graduated from HS WRT to Female & Not graduated from HS

```{r}
names(coef(fit3))
( fNoHS <- unname(sum(coef(fit3) * c(1, 0, 0, 0)) - coef(fit3)[1]) )
( mNoHS <- unname(sum(coef(fit3) * c(1, 1, 0, 0)) - coef(fit3)[1]) )
```

### Column 2

Male & Graduated from HS WRT Female & Not graduated from HS

```{r}
names(coef(fit3))
( fHS <- unname(sum(coef(fit3) * c(1, 0, 1, 0)) - coef(fit3)[1]) )
( mHS <- unname(sum(coef(fit3) * c(1, 1, 1, 0)) - coef(fit3)[1]) )
```

### Column 3

Male & Graduated from College WRT Female & Not graduated from HS

```{r}
names(coef(fit3))
coef(fit3)
( fC <- unname(sum(coef(fit3) * c(1, 0, 0, 1)) - coef(fit3)[1]) )
( mC <- unname(sum(coef(fit3) * c(1, 1, 0, 1)) - coef(fit3)[1]) )
```

## The table

```{r}
nhanes3$Sex <- factor(nhanes3$Male, levels = 0:1, labels = c("Female", "Male"))
contrasts(nhanes3$Sex)
table(nhanes3$Sex)
table(nhanes3$Male)
```


```{r}
no_int <- matrix(
    c(fNoHS, mNoHS, fHS, mHS, fC, mC),
    nrow = 2, dimnames = list(levels(nhanes3$Sex), levels(nhanes3$Edu))
)
no_int
# we subtract effect of males minus females
`Male-Female` <- no_int[2, ] - no_int[1, ]
no_int <- rbind(no_int, `Male-Female`)
no_int
```

## with Interaction

### Creating a common reference group

First, we check the reference categories:

```{r}
contrasts(nhanes3$Edu_f)
contrasts(factor(nhanes3$Male))
```

### "Female and Not Graduate from HS" is the common reference group

```{r}
nhanes3$SexEdu <- with(nhanes3, interaction(Sex, Edu_f))
contrasts(nhanes3$SexEdu)
```

```{r}
fit4 <- glm(Hypertension ~ SexEdu, family = "binomial", data = nhanes3)
summary(fit4)
```

### Columns 4, 5, and 6

|             | NoHS | HS | College |
|-------------|------|:--:|---------|
| Female      |      |    |         |
| Male        |      |    |         |
| Male-Female |      |    |         |

```{r}
coefs <- coef(fit4)
coefs[1] <- 0
with_int <- matrix(coefs, nrow = 2,
    dimnames = list(levels(nhanes3$Sex), levels(nhanes3$Edu)))
# we subtract effect of males minus females
`Male-Female` <- with_int[2, ] - with_int[1, ]
with_int <- rbind(with_int, `Male-Female`)
with_int
```

# Final Table

```{r}
round(cbind(no_int, with_int),3)
```

## Additive vs Multiplicative interaction?

As a good epidemiologist, you should be reporting both.
This is an important EPI topic.

https://catalyst.harvard.edu/docs/biostatsseminar/VanderWeele2012.pdf


# Lab 9 Exercise (subgroup analysis with HELP RCT)

Using the HELP RCT data, fit models with following variables:

* Response Y: linkstatus (1=linked; 0=otherwise)
* Covariate X1: treat (1=treated; 0=control)
* Covariate X2: substance (alcohol/cocaine/heroin)

Model 1: without interaction

Model 2: with interaction (X1 and X2)

Populate the following table and upload knitted Word document.

Notes. The 'table' can be a matrix as seen in the above example. In the table
columns below, `_i` means the estimate from the interaction.

|                 | Alcohol | Cocaine | Heroin | Alcohol_i | Cocaine_i | Heroin_i |
|-----------------|---------|---------|--------|-----------|-----------|----------|
| Control         |         |         |        |           |           |          |
| Treated         |         |         |        |           |           |          |
| Treated-Control |         |         |        |           |           |          |

